
---
title: "Statistical Analysis: A Retrospective Study: Inclusion of Risk factors in Scoring System to detect pre-test probability in patients diagnosed with Acute Pulmonary Embolism"

subtitle: Author: Julie Lee 
subtitle: Advisors: : Girdhar Sahil MD, Baveja Swati 

date: today
format:
  html:
    theme: cosmo
    embed-resources: true
    number-sections: true
    toc: true
    toc-depth: 4
    toc-location: left
    code-fold: false
engine: knitr
knitr:
  opts_chunk: 
    fig.align: 'center'
    # fig.width: 6
    # fig.height: 4
    message: FALSE
    cache: false
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, cache = FALSE)
```

```{r}
library(readxl)
library(dplyr)
library(readr)
library(tibble) 
library(ggplot2) 
library(knitr) 
library(kableExtra) 
library(tidyr)
library(mice)
library(naniar)
library(randomForest)
library(purrr)
library(tibble)
```

# Part A: Data Wrangling
**Construction of the merged_data Dataset ** 

```{r}
full_data <- "~/Documents/Desert Valley Research /desert_valley.xlsx"
cases <- read_excel(full_data, sheet = 1)
controls <- read_excel(full_data, sheet = 2)
```


**New Variables: ** Three new variables were derived to enhance interpretability: systolic, diastolic, and bmi_marker. The systolic and diastolic variables were extracted by decomposing the original blood_pressure variable, which contained combined systolic/diastolic readings. Additionally, bmi_marker was created by categorizing the continuous bmi variable into six clinically meaningful BMI categories. 

```{r}
cases$heart_rate_geneva <- ifelse(
  cases$heart_rate >= 95, 3,
  ifelse(cases$heart_rate >= 75 & cases$heart_rate <= 94, 2, 1)
)

controls$heart_rate_geneva <- ifelse(
  controls$heart_rate >= 95, 3,
  ifelse(controls$heart_rate >= 75 & controls$heart_rate <= 94, 2, 1)
) 

bp_split_cases <- strsplit(cases$blood_pressure, "/")
cases$systolic <- as.numeric(sapply(bp_split_cases, `[`, 1))
cases$diastolic <- as.numeric(sapply(bp_split_cases, `[`, 2))

bp_split_controls <- strsplit(controls$blood_pressure, "/")
controls$systolic <- as.numeric(sapply(bp_split_controls, `[`, 1))
controls$diastolic <- as.numeric(sapply(bp_split_controls, `[`, 2))

cases$blood_pressure <- NULL
controls$blood_pressure <- NULL
```

```{r}
cases$bmi_marker <- cut(
  cases$bmi,
  breaks = c(-Inf, 18.5, 24.9, 29.9, 34.9, 39.9, Inf),
  labels = c(
    "Underweight", 
    "Normal", 
    "Overweight", 
    "Obesity Class 1", 
    "Obesity Class 2", 
    "Obesity Class 3"
  ),
  right = TRUE
)

controls$bmi_marker <- cut(
  controls$bmi,
  breaks = c(-Inf, 18.5, 24.9, 29.9, 34.9, 39.9, Inf),
  labels = c(
    "Underweight", 
    "Normal", 
    "Overweight", 
    "Obesity Class 1", 
    "Obesity Class 2", 
    "Obesity Class 3"
  ),
  right = TRUE
)
```

```{r}
cases$age_group <- cut(
  cases$age,
  breaks = c(18, 30, 40, 50, 60, 70, 80, 100),
  labels = c("19–30", "31–40", "41–50", "51–60", "61–70", "71–80", "81+"),
  right = TRUE,
  include.lowest = TRUE
)

controls$age_group <- cut(
  controls$age,
  breaks = c(18, 30, 40, 50, 60, 70, 80, 100),
  labels = c("19–30", "31–40", "41–50", "51–60", "61–70", "71–80", "81+"),
  right = TRUE,
  include.lowest = TRUE
)
```

**Handling Labeling for Binary Variables: ** For all binary (two-level factor) variables in the original dataset, data entries were standardized for consistency. Values coded as “3” were recoded as NA to indicate missingness, while values coded as “2” were recoded as 0. As a result, all binary variables follow a unified labeling convention: 1 = Yes, 0 = No, and NA = Missing.

```{r}
clean_binary_variable <- function(df, varname) {
  df[[varname]][df[[varname]] == 3] <- NA
  df[[varname]] <- factor(ifelse(df[[varname]] == 2, 0, df[[varname]]))
  return(df)
}

binary_vars <- c(
  "gender", "smoking_history", "copd_history", "emphysema_ct", "copd_exacerbation",
  "positive_d_dimer", "covid", "elevated_bnp", "elevated_troponin",
  "dvt_in_lower_extremity", "rvsp_based_age", "rv_hypokines_on_echo",
  "pulmonary_hypertension", "rh_thrombus", "rh_strain", "rv_lv_abnormality",
  "tricuspid_regurgitation", "rv_edd_abnormality", "septal_bowing", 
  "clinical_symptoms_dvt", "pe_more_likely", "heart_rate_100", 
  "immobilization_surgery", "dvt_pe", "haemoptysis", "malignancy",   
  "surgery_fracture", "active_malignancy", "unilateral_lower_limb_pain", 
  "pain_lower_limb_edema", "acute_resp"
)

for (var in binary_vars) {
  cases <- clean_binary_variable(cases, var)
}

for (var in binary_vars) {
  controls <- clean_binary_variable(controls, var)
}
```


**Handling Numerical Variables with Limited or Censored Data: ** Variables such as d_dimer_value, bnp_levels, troponin_levels, troponin_levels_hs, and map_value contained values reported with inequality signs (ex: “<0.015”). To standardize these entries for quantitative analysis, we parsed and retained only the numeric portion—interpreting values like “<0.015” as 0.015. 

```{r}
numeric_vars <- c("d_dimer_value", "bnp_levels", "troponin_levels", "troponin_levels_hs") 

cases[numeric_vars] <- lapply(cases[numeric_vars], parse_number)
controls[numeric_vars] <- lapply(controls[numeric_vars], parse_number)
```

```{r}
cases$map_value <- parse_number(trimws(cases$map_value))
controls$map_value <- parse_number(trimws(controls$map_value))
```

**Variable Type Conversion:** Each variable was converted to its appropriate data type to ensure accurate analysis. Continuous variables were retained or converted as numeric types, while categorical variables, particularly binary indicators, were converted into factors with consistent labeling (0 = No, 1 = Yes). 

```{r}
convert_variables <- function(df) {
  df <- df %>%
    mutate(
      pe_status = factor(pe_status), 
      age = as.numeric(age),
      gender = factor(gender), 
      race = factor(race), 
      bmi = as.numeric(bmi),
      bmi_marker = as.factor(bmi_marker), 

      smoking_history = factor(smoking_history),
      copd_history = factor(copd_history),
      emphysema_ct = factor(emphysema_ct),
      copd_exacerbation = factor(copd_exacerbation), 
      
      positive_d_dimer = factor(positive_d_dimer),
      d_dimer_value = as.numeric(d_dimer_value),
      
      covid = factor(covid),
      elevated_bnp = factor(elevated_bnp),
      bnp_levels = as.numeric(bnp_levels),
      elevated_troponin = factor(elevated_troponin),
      troponin_levels = as.numeric(troponin_levels),
      troponin_levels_hs = as.numeric(troponin_levels_hs),
      dvt_in_lower_extremity = factor(dvt_in_lower_extremity),
    
      rvsp_based_age = factor(rvsp_based_age),
      echo_rvsp_value = as.numeric(echo_rvsp_value),
      rv_hypokines_on_echo = factor(rv_hypokines_on_echo),
      pulmonary_hypertension = factor(pulmonary_hypertension),
      rh_thrombus = factor(rh_thrombus),
      rh_strain = factor(rh_strain),
      rv_to_lv_ratio = as.numeric(rv_to_lv_ratio),
      rv_lv_abnormality = factor(rv_lv_abnormality),
      
      tricuspid_regurgitation = factor(tricuspid_regurgitation),
      rv_edd_cm = as.numeric(rv_edd_cm),
      rv_edd_abnormality = factor(rv_edd_abnormality),
      septal_bowing = factor(septal_bowing),

      clinical_symptoms_dvt = factor(clinical_symptoms_dvt),
      pe_more_likely = factor(pe_more_likely),
      heart_rate_100 = factor(heart_rate_100),
      heart_rate = as.numeric(heart_rate),
      immobilization_surgery = factor(immobilization_surgery),
      dvt_pe = factor(dvt_pe),
      haemoptysis  = factor(haemoptysis),
      malignancy = factor(malignancy),

      surgery_fracture = factor(surgery_fracture),
      active_malignancy = factor(active_malignancy),
      unilateral_lower_limb_pain = factor(unilateral_lower_limb_pain),
      pain_lower_limb_edema = factor(pain_lower_limb_edema), 
      heart_rate_geneva = factor(heart_rate_geneva),
      
      well_score = as.numeric(well_score), 
      revised_geneva_score = as.numeric(revised_geneva_score), 
      admission_year = factor(admission_year), 
      map_value = as.numeric(map_value), 
      acute_resp = factor(acute_resp),
      diastolic = as.numeric(diastolic),
      systolic = as.numeric(systolic), 
      age_group = as.factor(age_group)
  
    )
  
  return(df)
}

cases <- convert_variables(cases)
controls <- convert_variables(controls)
```


**Merging the Cases and Controls Data: ** The final analytic dataset, named merged_data, comprises 366 observations across 50 variables. This dataset integrates individuals from both the case and control groups: 176 individuals diagnosed with pulmonary embolism (PE) and 190 individuals without PE. The outcome variable, pe_status, indicates case-control status, where a value of 1 denotes a case (PE present) and 0 denotes a control (PE absent).

```{r}
merged_data <- bind_rows(cases, controls) 
merged_data <- merged_data %>%
  select(-c(troponin_levels, troponin_levels_hs))
```

```{r}
str(merged_data)
```

# Part B: Cleaning Data: Dealing with Missing Values 

For the 52 variables presented, let us run a summary to see what percentage of the data are present in each variable: 

```{r}
missing_summary <- merged_data %>%
  summarise(across(everything(), ~mean(is.na(.)) * 100)) %>%
  pivot_longer(everything(), names_to = "Variable", values_to = "Percent_Missing") %>%
  filter(Percent_Missing > 0) %>%
  mutate(Category = case_when(
    Percent_Missing < 5 ~ "<5% Missing",
    Percent_Missing >= 5 & Percent_Missing <= 20 ~ "5–20% Missing",
    Percent_Missing > 20 ~ ">20% Missing"
  )) %>%
  arrange(desc(Percent_Missing))

minimal_missing <- missing_summary %>% filter(Category == "<5% Missing")
moderate_missing <- missing_summary %>% filter(Category == "5–20% Missing")
high_missing <- missing_summary %>% filter(Category == ">20% Missing")


minimal_missing %>%
  select(Variable, Percent_Missing) %>%
  kable("html", caption = "Less than 5% Missing") %>%
  kable_styling(full_width = FALSE, bootstrap_options = c("striped", "hover"))

moderate_missing %>%
  select(Variable, Percent_Missing) %>%
  kable("html", caption = "5–20% Missing") %>%
  kable_styling(full_width = FALSE, bootstrap_options = c("striped", "hover"))

high_missing %>%
  select(Variable, Percent_Missing) %>%
  kable("html", caption = "Greater than 20% Missing") %>%
  kable_styling(full_width = FALSE, bootstrap_options = c("striped", "hover"))
```

**Handling Missingess of Variables with <5% Missing Data ** 

For variables with less than 5% missingness, it is important to determine the underlying mechanism of missing data (MCAR, MAR, or NMAR) in order to select the appropriate method for handling it. Specifically, we assess whether the probability of missingness is independent of both observed and unobserved variables, which is a condition known as Missing Completely at Random (MCAR). To evaluate this, we applied Little’s MCAR test, which statistically compares means and covariances across groups defined by patterns of missing data. This test uses a chi-squared statistic to assess whether there are significant differences between groups with and without missing data.

In our analysis, the test returned a χ² statistic of 21.97 with 33 degrees of freedom and a p-value of 0.928. Since this p-value is well above the standard threshold of 0.05, we fail to reject the null hypothesis, indicating no significant differences between the observed data patterns. We therefore conclude that the data for the selected variables is MCAR. Furthermore, the Fisher’s Exact Test further confirms that there is no significant association between missingness in the subset of predictor variables and the outcome variable, pe_status, as indicated by a large p-value of 0.3221. This supports the conclusion that the missingness in this subset is independent of PE diagnosis and is therefore likely Missing Completely at Random (MCAR).

```{r}
vars_under_5 <- c("smoking_history", "emphysema_ct", "copd_history", "copd_exacerbation", 
                  "clinical_symptoms_dvt", "immobilization_surgery", "dvt_pe", 
                  "haemoptysis", "malignancy", "surgery_fracture", "active_malignancy", 
                  "unilateral_lower_limb_pain", "pain_lower_limb_edema", "well_score", 
                  "revised_geneva_score", "pe_more_likely")

data_subset<- merged_data[, vars_under_5]
mcar_test(data_subset)
```


```{r}
merged_data$any_missing_subset <- apply(is.na(data_subset), 1, any)
fisher.test(table(merged_data$any_missing_subset, merged_data$pe_status))
```

```{r}
merged_data <- merged_data %>%
  select(-c(any_missing_subset))
```

Because the missingness is determined to be completely at random (MCAR) and affects less than 5% of the data for each variable, we can apply either simple imputation or complete case analysis without introducing bias. We began by performing complete case analysis, removing all rows with missing values for the selected variables. This resulted in the exclusion of 9 rows, accounting for only 2.46% of the entire dataset. Such minimal data loss is unlikely to affect statistical power or validity, especially since 97.5% of the original sample is retained. 

```{r}
n_complete <- sum(complete.cases(data_subset))
n_total <- nrow(data_subset)
n_dropped <- n_total - n_complete

cat("Number of rows that would be dropped:", n_dropped, "\n")
cat("Percentage of data dropped:", round((n_dropped / n_total) * 100, 2), "%\n")
```

To ensure the robustness of our approach, we conducted a sensitivity analysis by comparing results from complete case analysis to those obtained using simple imputation. In the imputation method, missing values for categorical variables were replaced with the mode of each respective variable, while missing values for numerical variables were substituted with their corresponding mean. Logistic regression models were then fit on both the complete case and imputed datasets, and their AIC values were compared. The model based on complete case analysis yielded an AIC of 348.99, whereas the model based on the imputed dataset had an AIC of 353.70, suggesting that the complete case model provided a slightly better fit on the data that predicts the outcome variable, pe_status. Based on this comparison, we finalized our dataset using the complete case approach, resulting in the removal of 9 rows and retaining a total of 357 observations.

```{r}
data_complete_case <- na.omit(data_subset) 
get_mode <- function(x) {
  ux <- unique(x[!is.na(x)])
  ux[which.max(tabulate(match(x, ux)))]
}

data_imputed <- data_subset %>%
  mutate(across(where(is.numeric), ~ ifelse(is.na(.), mean(., na.rm = TRUE), .))) %>%
  mutate(across(where(is.factor), ~ ifelse(is.na(.), get_mode(.), .)))

data_complete_case$pe_status <- merged_data$pe_status[complete.cases(data_subset)]
data_imputed$pe_status <- merged_data$pe_status

model_cc <- glm(pe_status ~ ., data = data_complete_case, family = binomial)
model_imp <- glm(pe_status ~ ., data = data_imputed, family = binomial)

aic_values <- data.frame(
  Method = c("Complete Case Analysis", "Simple Imputation"),
  AIC = c(AIC(model_cc), AIC(model_imp))
)

aic_values %>%
  kable("html", caption = "Comparison of AIC Values: Complete Case vs Imputation") %>%
  kable_styling(full_width = FALSE, bootstrap_options = c("striped", "hover"))

merged_data <- merged_data[complete.cases(merged_data[, vars_under_5]), ]
```

**Handling Missingess of Variables with 5-20 percent missingness ** 

The only variable with moderate missingness (between 5–20%) is elevated_troponin, with 7.28% of values missing. To assess the nature of this missingness, we examine whether it is associated with the outcome variable, pe_status (pulmonary Embolism diagnosis).

A cross-tabulation shows that among individuals diagnosed with PE, 170 had observed elevated_troponin values while 17 had missing values. Among those without PE, 161 had observed values and 9 had missing values. The similar proportions of missingness in both groups suggest that the absence of data is not systematically linked to PE status—supporting the assumption that the data may be Missing Completely At Random (MCAR).

We also examined the PE rate within each group. Among those with observed troponin values, the PE rate is 48.6%, whereas it is 34.6% among those with missing troponin values. To determine whether this difference is statistically significant, we conducted both a Chi-square test of independence and Little’s MCAR test.

```{r}
troponin_pe_table <- table(is.na(merged_data$elevated_troponin), merged_data$pe_status)

troponin_pe_df <- as.data.frame.matrix(troponin_pe_table)
rownames(troponin_pe_df) <- c("Observed (Not Missing)", "Missing")
troponin_pe_df <- tibble::rownames_to_column(troponin_pe_df, var = "Troponin Missingness")
colnames(troponin_pe_df)[2:3] <- c("PE Absent (0)", "PE Present (1)")

troponin_pe_df %>%
  kable("html", caption = "Cross-tabulation of Troponin Missingness by PE Status") %>%
  kable_styling(bootstrap_options = c("striped", "hover", "condensed"), full_width = FALSE)
```

```{r}
merged_data %>%
  mutate(pe_status_num = as.numeric(as.character(pe_status)),
         troponin_missing = ifelse(is.na(elevated_troponin), "Missing", "Not Missing")) %>%
  group_by(troponin_missing) %>%
  summarise(pe_rate = mean(pe_status_num, na.rm = TRUE),
            n = n()) %>%
  kable("html", caption = "Pulmonary Embolism Rate by Elevated Troponin Missingness") %>%
  kable_styling(full_width = FALSE, bootstrap_options = c("striped", "hover"))
```

The Chi-square test yielded a p-value of 0.24, indicating no significant association between missingness in elevated_troponin and pe_status. The Little’s MCAR test also returned a non-significant p-value (0.168), reinforcing the result.

```{r}
chisq.test(table(is.na(merged_data$elevated_troponin), merged_data$pe_status))
mcar_test(merged_data[,c("elevated_troponin","pe_status")]) 
```

These findings suggest that the missingness in elevated_troponin is not systematically related to observed data, and thus can be considered MCAR. However, it is reasonable to proceed under the MAR (Missing At Random) framework for multiple imputation rather than complete case analysis. A complete case approach would result in the exclusion of 26 observations, which could reduce statistical power and introduce bias, especially if the missingness is related to the outcome variable, PE status. In order to handle the missing values in elevated_troponin, we applied Multiple Imputation by Chained Equations (MICE) using logistic regression imputation. 

The imputation model included the outcome variable pe_status and additional covariates to preserve the variability and account for potential relationships between elevated_troponin and other predictors. To identify relevant covariates for imputation, we trained a Random Forest classifier on the observed data and ranked the predictors based on their Mean Decrease in Accuracy (a measure of variable importance.) From variables with no missingness, we selected the top four predictors: acute_resp, map_value, bmi, and heart_rate. These were chosen based on their predictive strength. Additionally, we included age (biologically relevant covaraite) and pe_status (the outcome variable)  These covariates, together with elevated_troponin, were included in the MICE procedure with 5 imputed datasets and 10 iterations, using the first completed dataset to impute the missing values.

```{r}
no_missing_vars <- names(merged_data)[colSums(is.na(merged_data)) == 0]
merged_data_no_missing <- merged_data[, no_missing_vars]
```

```{r}
formula_str <- paste(
  "elevated_troponin ~", 
  paste(no_missing_vars[no_missing_vars != "elevated_troponin"], collapse = " + ")
)
rf_formula <- as.formula(formula_str)

rf_data <- merged_data %>% filter(!is.na(elevated_troponin))

set.seed(123)
rf_model <- randomForest(
  formula = rf_formula,
  data = rf_data,
  importance = TRUE, 
  ntree = 500
)

var_importance <- importance(rf_model, type = 1) 
top_vars <- rownames(var_importance)[order(-var_importance[, 1])][1:10]

importance_df <- as.data.frame(importance(rf_model))
importance_df$Variable <- rownames(importance_df)

top10_vars <- importance_df %>%
  arrange(desc(MeanDecreaseAccuracy)) %>%
  select(Variable, MeanDecreaseAccuracy) %>%
  slice(1:10)

top10_vars %>%
  kable("html", caption = "Top 10 Predictors of Elevated Troponin (by Mean Decrease Accuracy)") %>%
  kable_styling(bootstrap_options = c("striped", "hover", "condensed"), full_width = FALSE)
```

From the specified list, all combinations of 3 to 6 predictors are iteratively evaluated to identify the optimal set for imputing missing values in the elevated_troponin variable using logistic regression via the mice package. For each combination, the script conducts multiple imputation, compares the proportions of observed versus imputed values, and fits logistic regression models before and after imputation to assess the effect on pe_status. A composite score is calculated for each set, with lower scores indicating better performance which reflects  minimal differences in class proportions, regression coefficients for elevated_troponin, and associated p-values between the complete-case and imputed models.


```{r}
top_predictors <- c("bmi", "acute_resp", "bmi_marker", "map_value", "heart_rate", 
                    "well_score", "heart_rate_100", "diastolic", 
                    "unilateral_lower_limb_pain", "active_malignancy")

outcome <- "elevated_troponin"

evaluate_combo <- function(predictors) {
  impute_vars <- c(outcome, "pe_status", predictors)
  impute_data <- merged_data[, impute_vars]

  imp <- tryCatch({
    mice(impute_data, method = "logreg", m = 5, maxit = 10, seed = 123, print = FALSE)
  }, error = function(e) return(NULL))
  if (is.null(imp)) return(NULL)

  complete_data <- complete(imp, 1)
  temp_data <- merged_data
  temp_data$imputed_tr <- complete_data[[outcome]]

  mask_obs <- !is.na(impute_data[[outcome]])
  observed <- temp_data[mask_obs, outcome, drop = TRUE]
  imputed <- temp_data[!mask_obs, "imputed_tr", drop = TRUE]

  prop_obs_1 <- mean(observed == 1, na.rm = TRUE)
  prop_obs_0 <- mean(observed == 0, na.rm = TRUE)
  prop_imp_1 <- mean(imputed == 1, na.rm = TRUE)
  prop_imp_0 <- mean(imputed == 0, na.rm = TRUE)

  prop_diff_1 <- abs(prop_obs_1 - prop_imp_1)
  prop_diff_0 <- abs(prop_obs_0 - prop_imp_0)

  model_complete <- tryCatch({
    glm(pe_status ~ get(outcome),
        data = temp_data[mask_obs, ],
        family = binomial)
  }, error = function(e) return(NULL))

  model_imputed <- tryCatch({
    glm(pe_status ~ imputed_tr,
        data = temp_data,
        family = binomial)
  }, error = function(e) return(NULL))

  if (is.null(model_complete) || is.null(model_imputed)) return(NULL)

  coef_complete <- tryCatch(coef(model_complete)[2], error = function(e) NA)
  coef_imputed <- tryCatch(coef(model_imputed)[2], error = function(e) NA)
  pval_complete <- tryCatch(summary(model_complete)$coefficients[2, 4], error = function(e) NA)
  pval_imputed <- tryCatch(summary(model_imputed)$coefficients[2, 4], error = function(e) NA)

  coef_diff <- abs(coef_complete - coef_imputed)
  pval_diff <- abs(pval_complete - pval_imputed)

  score <- prop_diff_1 + prop_diff_0 + coef_diff + pval_diff

  return(tibble(
    predictors = paste(predictors, collapse = ", "),
    prop_diff_1 = round(prop_diff_1, 3),
    prop_diff_0 = round(prop_diff_0, 3),
    coef_diff = round(coef_diff, 3),
    pval_diff = round(pval_diff, 3),
    score = round(score, 3)
  ))
}

combinations <- unlist(lapply(3:6, function(n) combn(top_predictors, n, simplify = FALSE)), recursive = FALSE)
safe_results <- map(combinations, safely(evaluate_combo))
successful_results <- keep(safe_results, ~ is.null(.x$error))
results <- map_dfr(successful_results, ~ as_tibble(.x$result)) %>%
  arrange(score)

print(head(results, 10))
```


```{r}
impute_vars <- c("pe_status", "elevated_troponin", "bmi", "acute_resp", "bmi_marker", "map_value", "well_score") 
impute_data <- merged_data[, impute_vars]
```

```{r, message=FALSE, warning=FALSE, results='hide'}
capture.output({
  imputed_data <- mice(
    impute_data,
    m = 5,
    method = "logreg",
    maxit = 10,
    seed = 123
  )
}, file = NULL)
```

```{r}
pre_imputation_data <- merged_data %>%
  filter(!is.na(elevated_troponin))
```

```{r}
completed_data <- complete(imputed_data, 1)
merged_data$elevated_troponin[is.na(merged_data$elevated_troponin)] <- 
  completed_data$elevated_troponin[is.na(merged_data$elevated_troponin)]
```


Sensitivity Analysis to Justify Multiple Imputation of elevated_troponin: The bar plot comparing observed and imputed values of elevated_troponin indicates that the proportions of levels (0 vs. 1) are reasonably consistent across both groups. This further suggests that the imputation preserved the original distribution of the data. 

```{r}
merged_data %>%
  mutate(trop_flag = ifelse(is.na(impute_data$elevated_troponin), "Imputed", "Observed")) %>%
  group_by(trop_flag, elevated_troponin) %>%
  summarise(n = n(), .groups = "drop") %>%
  group_by(trop_flag) %>%
  mutate(prop = n / sum(n)) %>%
  ggplot(aes(x = factor(elevated_troponin), y = prop, fill = trop_flag)) +
  geom_bar(stat = "identity", position = "dodge") +
  labs(title = "Proportion of Elevated Troponin (Observed vs Imputed)",
       x = "Elevated Troponin", y = "Proportion") +
  theme_minimal()
```

To assess whether imputation influenced model results, we fit two logistic regression models: one using the complete-case data (before imputation) and another using the full dataset after imputing missing elevated_troponin values. Both models included the same set of relevant predictors used in the imputation model. The results show that the interpretation of the model remains consistent before and after imputation: The elevated_troponin variable is not a statistically significant predictor of pulmonary embolism in either model. In addition, the coefficient for elevated_troponin_1 changed insignificantly after imputation (-0.143 to -0.157) which indicates that the imputation of values did not significantly change the variable's relationship with the outcome variable, pe_status. This supports the conclusion that multiple imputation did not introduce bias or meaningfully alter the relationships among variables. This validates the appropriateness of the imputation procedure.

```{r}
impute_vars <- c("elevated_troponin", "bmi", "acute_resp", "bmi_marker", "map_value", "well_score") 
logit_formula <- as.formula(paste("pe_status ~", paste(impute_vars, collapse = " + ")))

model_complete <- glm(logit_formula,
                      data = pre_imputation_data,
                      family = binomial)

model_imputed <- glm(logit_formula,
                     data = merged_data,
                     family = binomial)

summary_complete <- summary(model_complete)
summary_imputed <- summary(model_imputed)

logit_table <- data.frame(
  Coef_Complete = round(summary_complete$coefficients[, "Estimate"], 3),
  P_Complete = round(summary_complete$coefficients[, "Pr(>|z|)"], 3),
  Coef_Imputed = round(summary_imputed$coefficients[, "Estimate"], 3),
  P_Imputed = round(summary_imputed$coefficients[, "Pr(>|z|)"], 3)
)


logit_table_sorted <- logit_table[order(logit_table$P_Imputed), ]

logit_table_sorted %>%
  kable("html", caption = "Logistic Regression Comparison Between Pre and Post Imputated Data") %>%
  kable_styling(bootstrap_options = c("striped", "hover", "condensed"), full_width = FALSE)

```


**Handling Missingness of Variables with more than 20 percent missingness ** 

The variables with more than 80 percent missingness include: rv_to_lv_ratio, rv_lv_abnormality, rv_edd_cm, and rv_edd_abnormality. The best approach to handle such variables is to remove them from the dataset. Both simple and multiple imputation methods are highly unreliable to deal with such a large proportion of missingness and might create biased results. After removing these variables, we end up with 48 remaining variables. 

```{r}
merged_data <- merged_data %>%
  select(-c(rv_to_lv_ratio, rv_lv_abnormality, rv_edd_cm, rv_edd_abnormality))
```


**Variables: positive_d_dimer and d_dimer_value **

The variable d_dimer_value has 24.9 percent missing data while the variable positive_d_dimer also has 24.9 percent missing data. Based on the cross-tabulation output, among individuals diagnosed with pulmonary embolism (PE), 139 had observed values for d_dimer_levels (a numerical measure of a protein fragment produced when a blood clot breaks down), while 48 had missing entries. Similarly, among individuals without PE, 130 had observed values and 40 had missing values. The comparable distribution of missingness between the PE and non-PE groups suggests that the missing values in d_dimer_value are not systematically related to PE status. This pattern supports the assumption that the missingness may be Missing Completely at Random (MCAR).

```{r}
d_dimer_pe_table <- table(is.na(merged_data$d_dimer_value), merged_data$pe_status)
d_dimer_pe_df <- as.data.frame.matrix(d_dimer_pe_table)
rownames(d_dimer_pe_df) <- c("Observed (Not Missing)", "Missing")
d_dimer_pe_df <- tibble::rownames_to_column(d_dimer_pe_df, var = "d-Dimer Levels Missingness")
colnames(d_dimer_pe_df)[2:3] <- c("PE Absent (0)", "PE Present (1)")
d_dimer_pe_df %>%
  kable("html", caption = "Cross-tabulation of d-Dimer Levels Missingness by PE Status") %>%
  kable_styling(bootstrap_options = c("striped", "hover", "condensed"), full_width = FALSE)
```


Furthermore, the chi-square test of independence yields a large p-value (0.7298), indicating that there is a nonsignificant association between missingness in d_dimer_value and the outcome variable, pe_status. This further supports the conclusion that missingness in this variable may possibly be completely at random (MCAR). The Little's MCAR test also showcases a large p-value (p-value = 0.6395268), reinforcing the result. 

```{r}
chisq.test(table(is.na(merged_data$d_dimer_value), merged_data$pe_status))
mcar_test(merged_data[,c("d_dimer_value","pe_status")]) 
```

Logistic Regression: To investigate the missingness mechanism of the d_dimer_value variable, we conducted a logistic regression analysis to assess whether the probability of missingness was associated with other observed variables. We first created a binary indicator variable, missing_d_dimer_value where 1 indicates a missing value and 0 indicates an observed one. To avoid introducing bias or excessive row deletion during logistic regression, we restricted the analysis to a subset of variables with no missing data. The logistic regression results indicated that pe_status was not a significant predictor of missingness. However, variables such as well_score and admission_year predicted missingness in d_dimer_values. This suggests that while the missingness may appear close to Missing Completely at Random (MCAR), there is some dependence on observed variables, making the missingness mechanism more aligned with Missing at Random (MAR). Therefore, we proceed with multiple imputation rather than complete case analysis to retain statistical power and reduce bias.

```{r}
merged_data$missing_d_dimer_value<- as.numeric(is.na(merged_data$d_dimer_value))
```

```{r}
no_missing_vars <- names(merged_data)[colSums(is.na(merged_data)) == 0]
merged_data_no_missing <- merged_data[, no_missing_vars]
```

```{r}
excluded_vars <- c("d_dimer_value", "missing_d_dimer_value", "positive_d_dimer")
predictor_vars <- setdiff(names(merged_data_no_missing), excluded_vars)

formula_str <- paste("missing_d_dimer_value ~", paste(predictor_vars, collapse = " + "))
formula <- as.formula(formula_str)

missingness_model <- glm(formula, data = merged_data_no_missing, family = binomial)

significant_vars <- tidy(missingness_model) %>%
  filter(term != "(Intercept)", p.value < 0.1) %>%
  arrange(p.value)

significant_vars %>%
  mutate(p.value = signif(p.value, 3)) %>%
  kable("html", caption = "Significant Predictors of Missingness in d-dimer values") %>%
  kable_styling(bootstrap_options = c("striped", "hover", "condensed"), full_width = FALSE)
```

Imputation for d_dimer values: Missing values in d_dimer_values were imputed using Predictive Mean Matching (PMM) via the MICE algorithm. The predictor variables used in the imputation include both: (1) variables most strongly associated with missingness in d_dimer_values (as detailed in the table above). To address the skewed distribution of d-dimer values, the values of the variable were log-transformed using log1p() prior to imputation. PMM was then applied over 5 imputed datasets and 10 iterations. After imputation, the values were back-transformed to their original scale using expm1(). Finally, the binary variable positive_d_dimer was derived from the imputed d_dimer_value, coded as 1 for values greater than 0.5 ug/mL, and 0 otherwise.

```{r}
mandatory_predictors <- c(
  "age", "well_score", "race", "pe_more_likely", "admission_year", "gender", "revised_geneva_score", "pe_status", "heart_rate_geneva"
)

impute_predictors <- c(mandatory_predictors) 
```

```{r}
merged_data$log_d_dimer <- log1p(merged_data$d_dimer_value)
impute_data <- merged_data[, c("log_d_dimer", impute_predictors)]
```

```{r}
pred_matrix <- make.predictorMatrix(impute_data)
pred_matrix[,] <- 0 
pred_matrix["log_d_dimer", impute_predictors] <- 1
```

```{r, message=FALSE, warning=FALSE, results='hide'}
capture.output(
  imp <- mice(impute_data,
              method = "pmm", 
              predictorMatrix = pred_matrix,
              m = 5, 
              maxit = 10,
              seed = 123),
  file = NULL
)
```

```{r}
log_d_dimer_imputed <- complete(imp, 1)$log_d_dimer
merged_data$d_dimer_value <- expm1(log_d_dimer_imputed)
```

```{r}
merged_data$d_dimer_value <- as.numeric(merged_data$d_dimer_value)
merged_data$positive_d_dimer <- factor(
  ifelse(merged_data$d_dimer_value > 0.5, 1, 0),
  levels = c(0, 1)
)
```

Validating Imputation of d_dimer_values and positive_d_dimer: The summary table presents descriptive statistics for d_dimer_values by imputed_flag. The observed and imputed groups show similar central tendencies, with mean values of 5.20 (observed) and 5.45 (imputed), and median values of 1.88 and 1.79, respectively. As expected, the standard deviation is lower in the imputed group, likely due to both the smoothing effect of imputation and a smaller sample size. Additionally, results from the Welch two-sample t-test and Wilcoxon rank-sum test yield p-values greater than 0.05, indicating no statistically significant difference in the means or distributions of d_dimer_values between the observed and imputed groups. This conclusion is further supported by the density plot, which shows substantial overlap between the two distributions and no notable shift post-imputation. Taken together, these findings confirm that the imputation of d-dimer values was  successful and did not introduce meaningful bias.

```{r}
imputed_df <- complete(imp, 1)
imputed_df$imputed_flag <- ifelse(is.na(merged_data$log_d_dimer), "Imputed", "Observed")

imputed_df$d_dimer_value <- expm1(imputed_df$log_d_dimer)

summary_table <- imputed_df %>%
  mutate(imputed_flag = factor(imputed_flag, levels = c("Observed", "Imputed"))) %>%
  group_by(imputed_flag) %>%
  summarise(
    Mean = round(mean(d_dimer_value, na.rm = TRUE), 2),
    Median = round(median(d_dimer_value, na.rm = TRUE), 2),
    SD = round(sd(d_dimer_value, na.rm = TRUE), 2),
    N = n(),
    .groups = 'drop'
  )

summary_table %>%
  kable("html", caption = "Summary of d-dimer by Imputation Status") %>%
  kable_styling(bootstrap_options = c("striped", "hover", "condensed", "responsive"))
```

```{r}
t.test(d_dimer_value ~ imputed_flag, data = imputed_df)
wilcox.test(d_dimer_value ~ imputed_flag, data = imputed_df)
```

```{r}
ggplot(imputed_df, aes(x = d_dimer_value, fill = imputed_flag)) +
  geom_density(alpha = 0.4) +
  labs(
    title = "Distribution of Observed vs. Imputed d-dimer Levels",
    x = "BNP Levels",
    y = "Density",
    fill = "Value Type"
  ) +
  theme_minimal()
```

```{r}
merged_data <- merged_data %>% select(-log_d_dimer)
merged_data <- merged_data %>% select(-missing_d_dimer_value)
```


**Variables: bnp_levels and elevated_bnp ** 
The variable bnp_levels has 28.4 percent missing data while the variable elevated_bnp has 27.6 percent missing data. Based on the cross-tabulation output, among individuals diagnosed with pulmonary embolism (PE), 134 had observed values for bnp_levels (a numerical measure of B-type natriuretic peptide), while 53 had missing entries. Similarly, among individuals without PE, 123 had observed values and 47 had missing values. The comparable distribution of missingness between the PE and non-PE groups suggests that the missing values in bnp_levels are not systematically related to PE status. This pattern supports the assumption that the missingness may be Missing Completely at Random (MCAR).

```{r}
bnp_pe_table <- table(is.na(merged_data$bnp_levels), merged_data$pe_status)
bnp_pe_df <- as.data.frame.matrix(bnp_pe_table)
rownames(bnp_pe_df) <- c("Observed (Not Missing)", "Missing")
bnp_pe_df <- tibble::rownames_to_column(bnp_pe_df, var = "BNP Levels Missingness")
colnames(bnp_pe_df)[2:3] <- c("PE Absent (0)", "PE Present (1)")
bnp_pe_df %>%
  kable("html", caption = "Cross-tabulation of BNP Levels Missingness by PE Status") %>%
  kable_styling(bootstrap_options = c("striped", "hover", "condensed"), full_width = FALSE)
```

Furthermore, the chi-square test of independence yielded a small p-value (0.9776), indicating a nonsignificant association between missingness in bnp_levels and the outcome variable, pe_status. This further supports the conclusion that missingness in this variable is completely at random (MCAR). The Little's MCAR test also showcases a large p-value (p-value = 0.8838519), confirming that missingess is MCAR and that the chi-square test results are valid.

```{r}
chisq.test(table(is.na(merged_data$bnp_levels), merged_data$pe_status))
mcar_test(merged_data[,c("bnp_levels","pe_status")]) 
```

Logistic Regression: To investigate the missingness mechanism of the bnp_levels variable, we conducted a logistic regression analysis to assess whether the probability of missingness was associated with other observed variables. We first created a binary indicator variable, missing_bnp_levels, where 1 indicates a missing value and 0 indicates an observed one. To avoid introducing bias or excessive row deletion during logistic regression, we restricted the analysis to a subset of variables with no missing data. The logistic regression results indicated that pe_status was not a significant predictor of missingness. However, variables such as age and admission_year2020 weakly predicted missingness in bnp_levels. This suggests that while the missingness may appear close to Missing Completely at Random (MCAR), there is some dependence on observed variables, making the missingness mechanism more aligned with Missing at Random (MAR). Therefore, we proceed with multiple imputation rather than complete case analysis to retain statistical power and reduce bias.

```{r}
merged_data$missing_bnp_levels<- as.numeric(is.na(merged_data$bnp_levels))
```

```{r}
no_missing_vars <- names(merged_data)[colSums(is.na(merged_data)) == 0]
merged_data_no_missing <- merged_data[, no_missing_vars]
```

```{r}
excluded_vars <- c("bnp_levels", "missing_bnp_levels", "elevated_bnp")
predictor_vars <- setdiff(names(merged_data_no_missing), excluded_vars)

formula_str <- paste("missing_bnp_levels ~", paste(predictor_vars, collapse = " + "))
formula <- as.formula(formula_str)

missingness_model <- glm(formula, data = merged_data_no_missing, family = binomial)

significant_vars <- tidy(missingness_model) %>%
  filter(term != "(Intercept)", p.value < 0.1) %>%
  arrange(p.value)

significant_vars %>%
  mutate(p.value = signif(p.value, 3)) %>%
  kable("html", caption = "Significant Predictors of Missingness in bnp_levels") %>%
  kable_styling(bootstrap_options = c("striped", "hover", "condensed"), full_width = FALSE)
```

Imputation for bnp_levels: Missing values in bnp_levels were imputed using Predictive Mean Matching (PMM) via the MICE algorithm. The predictor variables used in the imputation include both: (1) variables most strongly associated with missingness in bnp_levels (as detailed in the table above), and (2) clinically relevant predictors known to be associated with both the outcome variable pe_status and with BNP (B-type Natriuretic Peptide), a key biomarker used to diagnose and assess heart failure. To address the skewed distribution of BNP, bnp_levels was log-transformed using log1p() prior to imputation. PMM was then applied over 5 imputed datasets and 10 iterations. After imputation, the values were back-transformed to their original scale using expm1(). Finally, the binary variable elevated_bnp was derived from the imputed bnp_levels, coded as 1 for values greater than 100 pg/mL, and 0 otherwise.

```{r}
mandatory_predictors <- c(
  "age", "admission_year", "copd_history", "pe_status", "bmi_marker"
)
clinical_predictors <- c(
  "heart_rate", "systolic", "diastolic"
)
impute_predictors <- c(mandatory_predictors, clinical_predictors)
```

```{r}
merged_data$log_bnp <- log1p(merged_data$bnp_levels)
impute_data <- merged_data[, c("log_bnp", impute_predictors)]
```

```{r}
pred_matrix <- make.predictorMatrix(impute_data)
pred_matrix[,] <- 0 
pred_matrix["log_bnp", impute_predictors] <- 1
```

```{r, message=FALSE, warning=FALSE, results='hide'}
capture.output(
  imp <- mice(impute_data,
              method = "pmm", 
              predictorMatrix = pred_matrix,
              m = 5, 
              maxit = 10,
              seed = 123),
  file = NULL
)
```

```{r}
log_bnp_imputed <- complete(imp, 1)$log_bnp
merged_data$bnp_levels <- expm1(log_bnp_imputed)
```

```{r}
merged_data$bnp_levels <- as.numeric(merged_data$bnp_levels)
merged_data$elevated_bnp <- factor(
  ifelse(merged_data$bnp_levels > 100, 1, 0),
  levels = c(0, 1)
)
```

Validating Imputation of bnp_levels and elevated_bnp: The summary table presents descriptive statistics for bnp_levels by imputed_flag. The observed and imputed groups show similar central tendencies, with mean values of 262.84 (observed) and 251.37 (imputed), and median values of 70.99 and 66.01, respectively. As expected, the standard deviation is lower in the imputed group, likely due to both the smoothing effect of imputation and a smaller sample size. Additionally, results from the Welch two-sample t-test and Wilcoxon rank-sum test yield p-values greater than 0.05, indicating no statistically significant difference in the means or distributions of bnp_levels between the observed and imputed groups. This conclusion is further supported by the density plot, which shows substantial overlap between the two distributions and no notable shift post-imputation. Taken together, these findings confirm that the imputation of bnp_levels was successful and did not introduce meaningful bias.

```{r}
imputed_df <- complete(imp, 1)
imputed_df$imputed_flag <- ifelse(is.na(merged_data$log_bnp), "Imputed", "Observed")

imputed_df$bnp_levels <- expm1(imputed_df$log_bnp)

summary_table <- imputed_df %>%
  mutate(imputed_flag = factor(imputed_flag, levels = c("Observed", "Imputed"))) %>%
  group_by(imputed_flag) %>%
  summarise(
    Mean = round(mean(bnp_levels, na.rm = TRUE), 2),
    Median = round(median(bnp_levels, na.rm = TRUE), 2),
    SD = round(sd(bnp_levels, na.rm = TRUE), 2),
    N = n(),
    .groups = 'drop'
  )

summary_table %>%
  kable("html", caption = "Summary of BNP Levels by Imputation Status") %>%
  kable_styling(bootstrap_options = c("striped", "hover", "condensed", "responsive"))
```

```{r}
t.test(bnp_levels ~ imputed_flag, data = imputed_df)
wilcox.test(bnp_levels ~ imputed_flag, data = imputed_df)
```

```{r}
ggplot(imputed_df, aes(x = bnp_levels, fill = imputed_flag)) +
  geom_density(alpha = 0.4) +
  labs(
    title = "Distribution of Observed vs. Imputed BNP Levels",
    x = "BNP Levels",
    y = "Density",
    fill = "Value Type"
  ) +
  theme_minimal()
```

```{r}
merged_data <- merged_data %>% select(-log_bnp)
merged_data <- merged_data %>% select(-missing_bnp_levels)
```















**Variables: covid ** 

The variable "covid" has 31.96 percent of its data missing. To assess the nature of this missingness, we examine whether it is associated with the outcome variable, pe_status. A cross-tabulation shows that among individuals diagnosed with PE, 127 had observed covid values while 60 had missing values. Among those without PE, 115 had observed values and 55 had missing values. The similar proportions of missingness in both groups suggest that the absence of data is not systematically linked to PE status, supporting the assumption that the data may be Missing Completely At Random (MCAR). We also examined the PE rate within each group. Among those with observed covid values, the PE rate is 0.475 percent, whereas it is 0.478 percent among those with missing covid values. 

```{r}
covid_pe_table <- table(is.na(merged_data$covid), merged_data$pe_status)

covid_pe_df <- as.data.frame.matrix(covid_pe_table)
rownames(covid_pe_df) <- c("Observed (Not Missing)", "Missing")
covid_pe_df <- tibble::rownames_to_column(covid_pe_df, var = "COVID Missingness")
colnames(covid_pe_df)[2:3] <- c("PE Absent (0)", "PE Present (1)")

covid_pe_df %>%
  kable("html", caption = "Cross-tabulation of COVID Missingness by PE Status") %>%
  kable_styling(bootstrap_options = c("striped", "hover", "condensed"), full_width = FALSE)
```

```{r}
merged_data %>%
  mutate(pe_status_num = as.numeric(as.character(pe_status)),
         covid_missing = ifelse(is.na(covid), "Missing", "Not Missing")) %>%
  group_by(covid_missing) %>%
  summarise(pe_rate = mean(pe_status_num, na.rm = TRUE),
            n = n()) %>%
  kable("html", caption = "Pulmonary Embolism Rate by COVID Missingness") %>%
  kable_styling(full_width = FALSE, bootstrap_options = c("striped", "hover"))
```


These findings suggest that the missingness in "covid" is not systematically related to observed data, and thus can be considered MCAR. However, it is reasonable to proceed under the MAR (Missing At Random) framework for multiple imputation rather than complete case analysis. A complete case approach would result in the exclusion of 115 observations, which could reduce statistical power and introduce bias, especially if the missingness is related to the outcome variable, PE status. In order to handle the missing values in covid, we applied Multiple Imputation by Chained Equations (MICE) using logistic regression imputation.

The imputation model included the outcome variable pe_status and additional covariates to preserve the variability and account for potential relationships between covid and other predictors. To identify relevant covariates for imputation, we trained a Random Forest classifier on the observed data and ranked the predictors based on their Mean Decrease in Accuracy. From variables with no missingness, we selected top variables: map_value, diastolic, heart_rate, age_group, systolic, well_score, and copd_history. These were chosen based on their predictive strength. Additionally, we included pe_status (the outcome variable). These covariates, together with covid, were included in the MICE procedure with 10 imputed datasets and 30 iterations, using the first completed dataset to impute the missing values.

```{r}
no_missing_vars <- names(merged_data)[colSums(is.na(merged_data)) == 0]
merged_data_no_missing <- merged_data[, no_missing_vars]
```

```{r}
formula_str <- paste(
  "covid ~", 
  paste(no_missing_vars[no_missing_vars != "covid"], collapse = " + ")
)
rf_formula <- as.formula(formula_str)

rf_data <- merged_data %>% filter(!is.na(covid))

set.seed(123)
rf_model <- randomForest(
  formula = rf_formula,
  data = rf_data,
  importance = TRUE, 
  ntree = 500
)

var_importance <- importance(rf_model, type = 1) 
top_vars <- rownames(var_importance)[order(-var_importance[, 1])][1:10]

importance_df <- as.data.frame(importance(rf_model))
importance_df$Variable <- rownames(importance_df)

top10_vars <- importance_df %>%
  arrange(desc(MeanDecreaseAccuracy)) %>%
  select(Variable, MeanDecreaseAccuracy) %>%
  slice(1:10)

top10_vars %>%
  kable("html", caption = "Top 10 Predictors of COVID-19 (by Mean Decrease Accuracy)") %>%
  kable_styling(bootstrap_options = c("striped", "hover", "condensed"), full_width = FALSE)
```


The same iterative imputation procedure is applied to the variable covid to find the optimal set for imputing missing covid values. The code is essentially the same as iterative imputation performed for previous variables above. Additionally, we included pe_status (the outcome variable). These covariates, together with covid, were included in the MICE procedure with 5 imputed datasets and 10 iterations, using the first completed dataset to impute the missing values.


```{r}
impute_vars <- c("pe_status", "covid", "map_value", "diastolic", "heart_rate", "age", "elevated_troponin")
impute_data <- merged_data[, impute_vars]
```

```{r, message=FALSE, warning=FALSE, results='hide'}
capture.output({
  imputed_data <- mice(
    impute_data,
    m = 5,
    method = "logreg",
    maxit = 10,
    seed = 123
  )
}, file = NULL)
```

```{r}
pre_imputation_data <- merged_data %>%
  filter(!is.na(covid))
```

```{r}
completed_data <- complete(imputed_data, 1)
merged_data$covid[is.na(merged_data$covid)] <- 
  completed_data$covid[is.na(merged_data$covid)]
```

Sensitivity Analysis to Justify Multiple Imputation of covid: The bar plot comparing observed and imputed values of covid indicates that the proportions of levels (0 vs. 1) are reasonably consistent across both groups. This further suggests that the imputation preserved the original distribution of the data. 

```{r}
merged_data %>%
  mutate(covid_flag = ifelse(is.na(impute_data$covid), "Imputed", "Observed")) %>%
  group_by(covid_flag, covid) %>%
  summarise(n = n(), .groups = "drop") %>%
  group_by(covid_flag) %>%
  mutate(prop = n / sum(n)) %>%
  ggplot(aes(x = factor(covid), y = prop, fill = covid_flag)) +
  geom_bar(stat = "identity", position = "dodge") +
  labs(title = "Proportion of COVID-19 (Observed vs Imputed)",
       x = "COVID - 19", y = "Proportion") +
  theme_minimal()
```

To assess whether imputation influenced model results, we fit two logistic regression models: one using the complete-case data (before imputation) and another using the full dataset after imputing missing covid values. Both models included the same set of relevant predictors used in the imputation model. The results show that the interpretation of the model remains consistent before and after imputation: The covid variable is not a statistically significant predictor of pulmonary embolism in either model. In addition, the coefficient for covid changed insignificantly after imputation (-0.227 to -0.131) which indicates that the imputation of values did not significantly change the variable's relationship with the outcome variable, pe_status. This validates the appropriateness of the imputation procedure.

```{r}
impute_vars <- c("covid", "map_value", "diastolic", "heart_rate", "age", "elevated_troponin")
logit_formula <- as.formula(paste("pe_status ~", paste(impute_vars, collapse = " + ")))

model_complete <- glm(logit_formula,
                      data = pre_imputation_data,
                      family = binomial)

model_imputed <- glm(logit_formula,
                     data = merged_data,
                     family = binomial)

summary_complete <- summary(model_complete)
summary_imputed <- summary(model_imputed)

logit_table <- data.frame(
  Coef_Complete = round(summary_complete$coefficients[, "Estimate"], 3),
  P_Complete = round(summary_complete$coefficients[, "Pr(>|z|)"], 3),
  Coef_Imputed = round(summary_imputed$coefficients[, "Estimate"], 3),
  P_Imputed = round(summary_imputed$coefficients[, "Pr(>|z|)"], 3)
)


logit_table_sorted <- logit_table[order(logit_table$P_Imputed), ]

logit_table_sorted %>%
  kable("html", caption = "Logistic Regression Comparison Between Pre and Post Imputated Data") %>%
  kable_styling(bootstrap_options = c("striped", "hover", "condensed"), full_width = FALSE)
```


**Variable: dvt_in_lower_extremity ** 

The variable "dvt_in_lower_extremity" has 35.79 percent of its data missing. To assess the nature of this missingness, we examine whether it is associated with the outcome variable, pe_status. A cross-tabulation shows that among individuals diagnosed with PE, 83 had observed dvt values while 104 had missing values. Among those without PE, 146 had observed values and 24 had missing values. The significant difference in proportions of missingness in both groups suggest that the absence of data is systematically linked to PE status, supporting the assumption that the data may not be Missing Completely At Random (MCAR). We also examined the PE rate within each group. Among those with observed dvt values, the PE rate is 63.75 percent, whereas it is 18.75 percent among those with missing dvt values. 

```{r}
dvt_in_lower_extremity_pe_table <- table(is.na(merged_data$dvt_in_lower_extremity), merged_data$pe_status)

dvt_in_lower_extremity_pe_df <- as.data.frame.matrix(dvt_in_lower_extremity_pe_table)
rownames(dvt_in_lower_extremity_pe_df) <- c("Observed (Not Missing)", "Missing")
dvt_in_lower_extremity_pe_df <- tibble::rownames_to_column(dvt_in_lower_extremity_pe_df, var = "Deep Vein Thrombosis in Lower Extremity Missingness")
colnames(dvt_in_lower_extremity_pe_df)[2:3] <- c("PE Absent (0)", "PE Present (1)")

dvt_in_lower_extremity_pe_df %>%
  kable("html", caption = "Cross-tabulation of Deep Vein Thrombosis in Lower Extremity Missingness by PE Status") %>%
  kable_styling(bootstrap_options = c("striped", "hover", "condensed"), full_width = FALSE)
```

```{r}
merged_data %>%
  mutate(pe_status_num = as.numeric(as.character(pe_status)),
         dvt_in_lower_extremity_missing = ifelse(is.na(dvt_in_lower_extremity), "Missing", "Not Missing")) %>%
  group_by(dvt_in_lower_extremity_missing) %>%
  summarise(pe_rate = mean(pe_status_num, na.rm = TRUE),
            n = n()) %>%
  kable("html", caption = "Pulmonary Embolism Rate by DVT Missingness") %>%
  kable_styling(full_width = FALSE, bootstrap_options = c("striped", "hover"))
```


These findings suggest that the missingness in "dvt_in_lower_extremity" is systematically related to observed data, and thus can be considered MAR. Multiple imputation will conducted rather than complete case analysis. A complete case approach would result in the exclusion of 128 observations, which could reduce statistical power and introduce bias, especially if the missingness is related to the outcome variable, PE status. In order to handle the missing values in dvt, we applied Multiple Imputation by Chained Equations (MICE) using logistic regression imputation.

The imputation model included the outcome variable pe_status and additional covariates to preserve the variability and account for potential relationships between covid and other predictors. To identify relevant covariates for imputation, we trained a Random Forest classifier on the observed data and ranked the predictors based on their Mean Decrease in Accuracy. From variables with no missingness, we selected top variables which were chosen on predictive strength. 

```{r}
no_missing_vars <- names(merged_data)[colSums(is.na(merged_data)) == 0]
merged_data_no_missing <- merged_data[, no_missing_vars]
```

```{r}
formula_str <- paste(
  "dvt_in_lower_extremity ~", 
  paste(no_missing_vars[no_missing_vars != "dvt_in_lower_extremity"], collapse = " + ")
)
rf_formula <- as.formula(formula_str)

rf_data <- merged_data %>% filter(!is.na(dvt_in_lower_extremity))

set.seed(123)
rf_model <- randomForest(
  formula = rf_formula,
  data = rf_data,
  importance = TRUE, 
  ntree = 500
)

var_importance <- importance(rf_model, type = 1) 
top_vars <- rownames(var_importance)[order(-var_importance[, 1])][1:10]

importance_df <- as.data.frame(importance(rf_model))
importance_df$Variable <- rownames(importance_df)

top10_vars <- importance_df %>%
  arrange(desc(MeanDecreaseAccuracy)) %>%
  select(Variable, MeanDecreaseAccuracy) %>%
  slice(1:10)

top10_vars %>%
  kable("html", caption = "Top 10 Predictors of DVT (by Mean Decrease Accuracy)") %>%
  kable_styling(bootstrap_options = c("striped", "hover", "condensed"), full_width = FALSE)
```

The same iterative imputation procedure is applied to the variable dvt_in_lower_extremity to find the optimal set for imputing missing dvt values. The code is essentially the same as iterative imputation performed for previous variables above. Additionally, we included pe_status (the outcome variable). These covariates, together with dvt_in_lower_extremity, were included in the MICE procedure with 5 imputed datasets and 10 iterations, using the first completed dataset to impute the missing values. Code is not included as it is the same as the one from variables above. 


```{r}
impute_vars <- c("pe_status", "dvt_in_lower_extremity", "revised_geneva_score", "map_value", "dvt_pe")
impute_data <- merged_data[, impute_vars]
```

```{r, message=FALSE, warning=FALSE, results='hide'}
capture.output({
  imputed_data <- mice(
    impute_data,
    m = 5,
    method = "logreg",
    maxit = 10,
    seed = 123
  )
}, file = NULL)
```

```{r}
pre_imputation_data <- merged_data %>%
  filter(!is.na(dvt_in_lower_extremity))
```

```{r}
completed_data <- complete(imputed_data, 1)
merged_data$dvt_in_lower_extremity[is.na(merged_data$dvt_in_lower_extremity)] <- 
  completed_data$dvt_in_lower_extremity[is.na(merged_data$dvt_in_lower_extremity)]
```

Sensitivity Analysis to Justify Multiple Imputation of covid: The bar plot comparing observed and imputed values of dvt_in_lower_extremity indicates that the proportions of levels (0 vs. 1) are reasonably consistent across both groups. This further suggests that the imputation preserved the original distribution of the data. 

```{r}
merged_data %>%
  mutate(dvt_in_lower_extremity_flag = ifelse(is.na(impute_data$dvt_in_lower_extremity), "Imputed", "Observed")) %>%
  group_by(dvt_in_lower_extremity_flag, dvt_in_lower_extremity) %>%
  summarise(n = n(), .groups = "drop") %>%
  group_by(dvt_in_lower_extremity_flag) %>%
  mutate(prop = n / sum(n)) %>%
  ggplot(aes(x = factor(dvt_in_lower_extremity), y = prop, fill = dvt_in_lower_extremity_flag)) +
  geom_bar(stat = "identity", position = "dodge") +
  labs(title = "Proportion of DVT (Observed vs Imputed)",
       x = "DVT", y = "Proportion") +
  theme_minimal()
```

To assess whether imputation influenced model results, we fit two logistic regression models: one using the complete-case data (before imputation) and another using the full dataset after imputing missing dvt values. Both models included the same set of relevant predictors used in the imputation model. The results show that the interpretation of the model remains consistent before and after imputation: The dvt_in_lower_extremity variable is not a statistically significant predictor of pulmonary embolism in either model. In addition, the coefficient for dvt changed insignificantly after imputation (-1.476 to -1.067) which indicates that the imputation of values did not significantly change the variable's relationship with the outcome variable, pe_status. This validates the appropriateness of the imputation procedure.

```{r}
impute_vars <- c("dvt_in_lower_extremity", "revised_geneva_score", "map_value", "dvt_pe")
logit_formula <- as.formula(paste("pe_status ~", paste(impute_vars, collapse = " + ")))

model_complete <- glm(logit_formula,
                      data = pre_imputation_data,
                      family = binomial)

model_imputed <- glm(logit_formula,
                     data = merged_data,
                     family = binomial)

summary_complete <- summary(model_complete)
summary_imputed <- summary(model_imputed)

logit_table <- data.frame(
  Coef_Complete = round(summary_complete$coefficients[, "Estimate"], 3),
  P_Complete = round(summary_complete$coefficients[, "Pr(>|z|)"], 3),
  Coef_Imputed = round(summary_imputed$coefficients[, "Estimate"], 3),
  P_Imputed = round(summary_imputed$coefficients[, "Pr(>|z|)"], 3)
)


logit_table_sorted <- logit_table[order(logit_table$P_Imputed), ]

logit_table_sorted %>%
  kable("html", caption = "Logistic Regression Comparison Between Pre and Post Imputated Data") %>%
  kable_styling(bootstrap_options = c("striped", "hover", "condensed"), full_width = FALSE)
```


**Variable: tricuspid_regurgitation ** 

The variable "tricuspid_regurgitation" has 39.33 percent of its data missing. To assess the nature of this missingness, we examine whether it is associated with the outcome variable, pe_status. A cross-tabulation shows that among individuals diagnosed with PE, 98 had observed TR values while 89 had missing values. Among those without PE, 118 had observed values and 52 had missing values. The significant difference in proportions of missingness in both groups suggest that the absence of data is systematically linked to PE status, supporting the assumption that the data may not be Missing Completely At Random (MCAR). We also examined the PE rate within each group. Among those with observed TR values, the PE rate is 54.6 percent, whereas it is 36.87 percent among those with missing TR values. 

```{r}
tricuspid_regurgitation_pe_table <- table(is.na(merged_data$tricuspid_regurgitation), merged_data$pe_status)

tricuspid_regurgitation_pe_df <- as.data.frame.matrix(tricuspid_regurgitation_pe_table)
rownames(tricuspid_regurgitation_pe_df) <- c("Observed (Not Missing)", "Missing")
tricuspid_regurgitation_pe_df <- tibble::rownames_to_column(tricuspid_regurgitation_pe_df, var = "Tricuspid Regurgitation Missingness")
colnames(tricuspid_regurgitation_pe_df)[2:3] <- c("PE Absent (0)", "PE Present (1)")

tricuspid_regurgitation_pe_df %>%
  kable("html", caption = "Cross-tabulation of Tricuspid Regurgitation Missingness by PE Status") %>%
  kable_styling(bootstrap_options = c("striped", "hover", "condensed"), full_width = FALSE)
```

```{r}
merged_data %>%
  mutate(pe_status_num = as.numeric(as.character(pe_status)),
         tricuspid_regurgitation_missing = ifelse(is.na(tricuspid_regurgitation), "Missing", "Not Missing")) %>%
  group_by(tricuspid_regurgitation_missing) %>%
  summarise(pe_rate = mean(pe_status_num, na.rm = TRUE),
            n = n()) %>%
  kable("html", caption = "Pulmonary Embolism Rate by Tricuspid Regurgitation Missingness") %>%
  kable_styling(full_width = FALSE, bootstrap_options = c("striped", "hover"))
```


These findings suggest that the missingness in "tricuspid_regurgitation" is systematically related to observed data, and thus can be considered MAR. Multiple imputation will conducted rather than complete case analysis. A complete case approach would result in the exclusion of 141 observations, which could reduce statistical power and introduce bias, especially if the missingness is related to the outcome variable, PE status. In order to handle the missing values in TR, we applied Multiple Imputation by Chained Equations (MICE) using logistic regression imputation.

The imputation model included the outcome variable pe_status and additional covariates to preserve the variability and account for potential relationships between TR and other predictors. To identify relevant covariates for imputation, we trained a Random Forest classifier on the observed data and ranked the predictors based on their Mean Decrease in Accuracy. From variables with no missingness, we selected top variables which were chosen on predictive strength. 

```{r}
no_missing_vars <- names(merged_data)[colSums(is.na(merged_data)) == 0]
merged_data_no_missing <- merged_data[, no_missing_vars]
```

```{r}
formula_str <- paste(
  "tricuspid_regurgitation ~", 
  paste(no_missing_vars[no_missing_vars != "tricuspid_regurgitation"], collapse = " + ")
)
rf_formula <- as.formula(formula_str)

rf_data <- merged_data %>% filter(!is.na(tricuspid_regurgitation))

set.seed(123)
rf_model <- randomForest(
  formula = rf_formula,
  data = rf_data,
  importance = TRUE, 
  ntree = 500
)

var_importance <- importance(rf_model, type = 1) 
top_vars <- rownames(var_importance)[order(-var_importance[, 1])][1:10]

importance_df <- as.data.frame(importance(rf_model))
importance_df$Variable <- rownames(importance_df)

top10_vars <- importance_df %>%
  arrange(desc(MeanDecreaseAccuracy)) %>%
  select(Variable, MeanDecreaseAccuracy) %>%
  slice(1:10)

top10_vars %>%
  kable("html", caption = "Top 10 Predictors of Tricuspid Regurgitation (by Mean Decrease Accuracy)") %>%
  kable_styling(bootstrap_options = c("striped", "hover", "condensed"), full_width = FALSE)
```

The same iterative imputation procedure is applied to the variable tricuspid_regurgitation to find the optimal set for imputing missing TR values. The code is essentially the same as iterative imputation performed for previous variables above. Additionally, we included pe_status (the outcome variable). These covariates, together with tricuspid_regurgitation, were included in the MICE procedure with 5 imputed datasets and 10 iterations, using the first completed dataset to impute the missing values. Code is not included as it is the same as the one from variables above. 

```{r}
impute_vars <- c("pe_status", "tricuspid_regurgitation", "bnp_levels", "age", "map_value", "heart_rate", "revised_geneva_score", "well_score") 	
impute_data <- merged_data[, impute_vars]
```

```{r, message=FALSE, warning=FALSE, results='hide'}
capture.output({
  imputed_data <- mice(
    impute_data,
    m = 5,
    method = "logreg",
    maxit = 10,
    seed = 123
  )
}, file = NULL)
```

```{r}
pre_imputation_data <- merged_data %>%
  filter(!is.na(tricuspid_regurgitation))
```

```{r}
completed_data <- complete(imputed_data, 1)
merged_data$tricuspid_regurgitation[is.na(merged_data$tricuspid_regurgitation)] <- 
  completed_data$tricuspid_regurgitation[is.na(merged_data$tricuspid_regurgitation)]
```

Sensitivity Analysis to Justify Multiple Imputation of tricuspid_regurgitation: The bar plot comparing observed and imputed values of tricuspid_regurgitation indicates that the proportions of levels (0 vs. 1) are reasonably consistent across both groups. This further suggests that the imputation preserved the original distribution of the data. 

```{r}
merged_data %>%
  mutate(tricuspid_regurgitation_flag = ifelse(is.na(impute_data$tricuspid_regurgitation), "Imputed", "Observed")) %>%
  group_by(tricuspid_regurgitation_flag, tricuspid_regurgitation) %>%
  summarise(n = n(), .groups = "drop") %>%
  group_by(tricuspid_regurgitation_flag) %>%
  mutate(prop = n / sum(n)) %>%
  ggplot(aes(x = factor(tricuspid_regurgitation), y = prop, fill = tricuspid_regurgitation_flag)) +
  geom_bar(stat = "identity", position = "dodge") +
  labs(title = "Proportion of TR (Observed vs Imputed)",
       x = "DVT", y = "Proportion") +
  theme_minimal()
```

To assess whether imputation influenced model results, we fit two logistic regression models: one using the complete-case data (before imputation) and another using the full dataset after imputing missing dvt values.The results show that the interpretation of the model remains consistent before and after imputation: The tricuspid_regurgitation variable is not a statistically significant predictor of pulmonary embolism in either model. In addition, the coefficient for TR changed insignificantly after imputation (-0.337 to -0.270) which indicates that the imputation of values did not significantly change the variable's relationship with the outcome variable, pe_status. This validates the appropriateness of the imputation procedure.

```{r}
impute_vars <- c("tricuspid_regurgitation", "bnp_levels", "age", "map_value", "heart_rate", "revised_geneva_score", "well_score") 	
logit_formula <- as.formula(paste("pe_status ~", paste(impute_vars, collapse = " + ")))

model_complete <- glm(logit_formula,
                      data = pre_imputation_data,
                      family = binomial)

model_imputed <- glm(logit_formula,
                     data = merged_data,
                     family = binomial)

summary_complete <- summary(model_complete)
summary_imputed <- summary(model_imputed)

logit_table <- data.frame(
  Coef_Complete = round(summary_complete$coefficients[, "Estimate"], 3),
  P_Complete = round(summary_complete$coefficients[, "Pr(>|z|)"], 3),
  Coef_Imputed = round(summary_imputed$coefficients[, "Estimate"], 3),
  P_Imputed = round(summary_imputed$coefficients[, "Pr(>|z|)"], 3)
)


logit_table_sorted <- logit_table[order(logit_table$P_Imputed), ]

logit_table_sorted %>%
  kable("html", caption = "Logistic Regression Comparison Between Pre and Post Imputated Data") %>%
  kable_styling(bootstrap_options = c("striped", "hover", "condensed"), full_width = FALSE)
```


**Variable: septal_bowing ** 

The variable "septal_bowing" has 39.617 percent of its data missing. To assess the nature of this missingness, we examine whether it is associated with the outcome variable, pe_status. A cross-tabulation shows that among individuals diagnosed with PE, 98 had observed SB values while 89 had missing values. Among those without PE, 117 had observed values and 53 had missing values. The significant difference in proportions of missingness in both groups suggest that the absence of data is systematically linked to PE status, supporting the assumption that the data may not be Missing Completely At Random (MCAR). We also examined the PE rate within each group. Among those with observed septal bowing values, the PE rate is 54.4 percent, whereas it is 37.32 percent among those with missing septal bowing values. 

```{r}
septal_bowing_pe_table <- table(is.na(merged_data$septal_bowing), merged_data$pe_status)

septal_bowing_pe_df <- as.data.frame.matrix(septal_bowing_pe_table)
rownames(septal_bowing_pe_df) <- c("Observed (Not Missing)", "Missing")
septal_bowing_pe_df <- tibble::rownames_to_column(septal_bowing_pe_df, var = "Septal Bowing Missingness")
colnames(septal_bowing_pe_df)[2:3] <- c("PE Absent (0)", "PE Present (1)")

septal_bowing_pe_df %>%
  kable("html", caption = "Cross-tabulation of Septal Bowing Missingness by PE Status") %>%
  kable_styling(bootstrap_options = c("striped", "hover", "condensed"), full_width = FALSE)
```

```{r}
merged_data %>%
  mutate(pe_status_num = as.numeric(as.character(pe_status)),
         septal_bowing_missing = ifelse(is.na(septal_bowing), "Missing", "Not Missing")) %>%
  group_by(septal_bowing_missing) %>%
  summarise(pe_rate = mean(pe_status_num, na.rm = TRUE),
            n = n()) %>%
  kable("html", caption = "Pulmonary Embolism Rate by Septal Bowing") %>%
  kable_styling(full_width = FALSE, bootstrap_options = c("striped", "hover"))
```


These findings suggest that the missingness in "septal_bowing" is systematically related to observed data, and thus can be considered MAR. Multiple imputation will conducted rather than complete case analysis. A complete case approach would result in the exclusion of 142 observations, which could reduce statistical power and introduce bias, especially if the missingness is related to the outcome variable, PE status. In order to handle the missing values in SB, we applied Multiple Imputation by Chained Equations (MICE) using logistic regression imputation.

The imputation model included the outcome variable pe_status and additional covariates to preserve the variability and account for potential relationships between septal bowing and other predictors. To identify relevant covariates for imputation, we trained a Random Forest classifier on the observed data and ranked the predictors based on their Mean Decrease in Accuracy. From variables with no missingness, we selected top variables which were chosen on predictive strength. 

```{r}
no_missing_vars <- names(merged_data)[colSums(is.na(merged_data)) == 0]
merged_data_no_missing <- merged_data[, no_missing_vars]
```

```{r}
formula_str <- paste(
  "septal_bowing ~", 
  paste(no_missing_vars[no_missing_vars != "septal_bowing"], collapse = " + ")
)
rf_formula <- as.formula(formula_str)

rf_data <- merged_data %>% filter(!is.na(septal_bowing))

set.seed(123)
rf_model <- randomForest(
  formula = rf_formula,
  data = rf_data,
  importance = TRUE, 
  ntree = 500
)

var_importance <- importance(rf_model, type = 1) 
top_vars <- rownames(var_importance)[order(-var_importance[, 1])][1:10]

importance_df <- as.data.frame(importance(rf_model))
importance_df$Variable <- rownames(importance_df)

top10_vars <- importance_df %>%
  arrange(desc(MeanDecreaseAccuracy)) %>%
  select(Variable, MeanDecreaseAccuracy) %>%
  slice(1:10)

top10_vars %>%
  kable("html", caption = "Top 10 Predictors of Septal Bowing (by Mean Decrease Accuracy)") %>%
  kable_styling(bootstrap_options = c("striped", "hover", "condensed"), full_width = FALSE)
```

The same iterative imputation procedure is applied to the variable septal_bowing to find the optimal set for imputing missing SB values. The code is essentially the same as iterative imputation performed for previous variables above. Additionally, we included pe_status (the outcome variable). These covariates, together with septal_bowing, were included in the MICE procedure with 5 imputed datasets and 10 iterations, using the first completed dataset to impute the missing values. Code is not included as it is the same as the one from variables above. 

```{r}
impute_vars <- c("pe_status", "septal_bowing", "revised_geneva_score", "heart_rate", "map_value") 
impute_data <- merged_data[, impute_vars]
```

```{r, message=FALSE, warning=FALSE, results='hide'}
capture.output({
  imputed_data <- mice(
    impute_data,
    m = 5,
    method = "logreg",
    maxit = 10,
    seed = 123
  )
}, file = NULL)
```

```{r}
pre_imputation_data <- merged_data %>%
  filter(!is.na(septal_bowing))
```

```{r}
completed_data <- complete(imputed_data, 1)
merged_data$septal_bowing[is.na(merged_data$septal_bowing)] <- 
  completed_data$septal_bowing[is.na(merged_data$septal_bowing)]
```

Sensitivity Analysis to Justify Multiple Imputation of tricuspid_regurgitation: The bar plot comparing observed and imputed values of septal_bowing indicates that the proportions of levels (0 vs. 1) are reasonably consistent across both groups. This further suggests that the imputation preserved the original distribution of the data. 

```{r}
merged_data %>%
  mutate(septal_bowing_flag = ifelse(is.na(impute_data$septal_bowing), "Imputed", "Observed")) %>%
  group_by(septal_bowing_flag, septal_bowing) %>%
  summarise(n = n(), .groups = "drop") %>%
  group_by(septal_bowing_flag) %>%
  mutate(prop = n / sum(n)) %>%
  ggplot(aes(x = factor(septal_bowing), y = prop, fill = septal_bowing_flag)) +
  geom_bar(stat = "identity", position = "dodge") +
  labs(title = "Proportion of Septal Bowing (Observed vs Imputed)",
       x = "Septal Bowing", y = "Proportion") +
  theme_minimal()
```

To assess whether imputation influenced model results, we fit two logistic regression models: one using the complete-case data (before imputation) and another using the full dataset after imputing missing septal bowing values. The results show that the interpretation of the model remains consistent before and after imputation: The septal_bowing variable is statistically significant predictor of pulmonary embolism in either model at a alpha value of 0.10. In addition, the coefficient for SB changed insignificantly after imputation (-1.542 to -1.491) which indicates that the imputation of values did not significantly change the variable's relationship with the outcome variable, pe_status. This validates the appropriateness of the imputation procedure.

```{r}
impute_vars <- c("septal_bowing", "revised_geneva_score", "heart_rate", "map_value") 

logit_formula <- as.formula(paste("pe_status ~", paste(impute_vars, collapse = " + ")))

model_complete <- glm(logit_formula,
                      data = pre_imputation_data,
                      family = binomial)

model_imputed <- glm(logit_formula,
                     data = merged_data,
                     family = binomial)

summary_complete <- summary(model_complete)
summary_imputed <- summary(model_imputed)

logit_table <- data.frame(
  Coef_Complete = round(summary_complete$coefficients[, "Estimate"], 3),
  P_Complete = round(summary_complete$coefficients[, "Pr(>|z|)"], 3),
  Coef_Imputed = round(summary_imputed$coefficients[, "Estimate"], 3),
  P_Imputed = round(summary_imputed$coefficients[, "Pr(>|z|)"], 3)
)


logit_table_sorted <- logit_table[order(logit_table$P_Imputed), ]

logit_table_sorted %>%
  kable("html", caption = "Logistic Regression Comparison Between Pre and Post Imputated Data") %>%
  kable_styling(bootstrap_options = c("striped", "hover", "condensed"), full_width = FALSE)
```


**Variable: rh_strain ** 

The variable "rh_strain" or Right Heart Strain has 39.617 percent of its data missing. To assess the nature of this missingness, we examine whether it is associated with the outcome variable, pe_status. A cross-tabulation shows that among individuals diagnosed with PE, 98 had observed rh_strain values while 89 had missing values. Among those without PE, 117 had observed values and 53 had missing values. The significant difference in proportions of missingness in both groups suggest that the absence of data is systematically linked to PE status, supporting the assumption that the data may not be Missing Completely At Random (MCAR). We also examined the PE rate within each group. Among those with observed right heart strain values, the PE rate is 54.4 percent, whereas it is 37.32 percent among those with missing right heart strain values. 

```{r}
rh_strain_pe_table <- table(is.na(merged_data$rh_strain), merged_data$pe_status)

rh_strain_pe_df <- as.data.frame.matrix(rh_strain_pe_table)
rownames(rh_strain_pe_df) <- c("Observed (Not Missing)", "Missing")
rh_strain_pe_df <- tibble::rownames_to_column(rh_strain_pe_df, var = "Right Heart Strain Missingness")
colnames(rh_strain_pe_df)[2:3] <- c("PE Absent (0)", "PE Present (1)")

rh_strain_pe_df %>%
  kable("html", caption = "Cross-tabulation of Right Heart Strain Missingness by PE Status") %>%
  kable_styling(bootstrap_options = c("striped", "hover", "condensed"), full_width = FALSE)
```

```{r}
merged_data %>%
  mutate(pe_status_num = as.numeric(as.character(pe_status)),
         rh_strain_missing = ifelse(is.na(rh_strain), "Missing", "Not Missing")) %>%
  group_by(rh_strain_missing) %>%
  summarise(pe_rate = mean(pe_status_num, na.rm = TRUE),
            n = n()) %>%
  kable("html", caption = "Pulmonary Embolism Rate by Right Heart Strain") %>%
  kable_styling(full_width = FALSE, bootstrap_options = c("striped", "hover"))
```


These findings suggest that the missingness in "rh_strain" is systematically related to observed data, and thus can be considered MAR. Multiple imputation will conducted rather than complete case analysis. A complete case approach would result in the exclusion of 142 observations, which could reduce statistical power and introduce bias, especially if the missingness is related to the outcome variable, PE status. In order to handle the missing values in SB, we applied Multiple Imputation by Chained Equations (MICE) using logistic regression imputation.

The imputation model included the outcome variable pe_status and additional covariates to preserve the variability and account for potential relationships between rh_strain and other predictors. To identify relevant covariates for imputation, we trained a Random Forest classifier on the observed data and ranked the predictors based on their Mean Decrease in Accuracy. From variables with no missingness, we selected top variables which were chosen on predictive strength. 

```{r}
no_missing_vars <- names(merged_data)[colSums(is.na(merged_data)) == 0]
merged_data_no_missing <- merged_data[, no_missing_vars]
```

```{r}
formula_str <- paste(
  "rh_strain ~", 
  paste(no_missing_vars[no_missing_vars != "rh_strain"], collapse = " + ")
)
rf_formula <- as.formula(formula_str)

rf_data <- merged_data %>% filter(!is.na(rh_strain))

set.seed(123)
rf_model <- randomForest(
  formula = rf_formula,
  data = rf_data,
  importance = TRUE, 
  ntree = 500
)

var_importance <- importance(rf_model, type = 1) 
top_vars <- rownames(var_importance)[order(-var_importance[, 1])][1:10]

importance_df <- as.data.frame(importance(rf_model))
importance_df$Variable <- rownames(importance_df)

top10_vars <- importance_df %>%
  arrange(desc(MeanDecreaseAccuracy)) %>%
  select(Variable, MeanDecreaseAccuracy) %>%
  slice(1:10)

top10_vars %>%
  kable("html", caption = "Top 10 Predictors of Right Heart Strain (by Mean Decrease Accuracy)") %>%
  kable_styling(bootstrap_options = c("striped", "hover", "condensed"), full_width = FALSE)
```

The same iterative imputation procedure is applied to the variable rh_strain to find the optimal set for imputing missing values. The code is essentially the same as iterative imputation performed for previous variables above. Additionally, we included pe_status (the outcome variable). These covariates, together with rh_strain were included in the MICE procedure with 5 imputed datasets and 10 iterations, using the first completed dataset to impute the missing values. Code is not included as it is the same as the one from variables above. 

```{r}
impute_vars <- c("pe_status", "rh_strain", "systolic", "map_value", "revised_geneva_score", "well_score") 
impute_data <- merged_data[, impute_vars]
```

```{r, message=FALSE, warning=FALSE, results='hide'}
capture.output({
  imputed_data <- mice(
    impute_data,
    m = 5,
    method = "logreg",
    maxit = 10,
    seed = 123
  )
}, file = NULL)
```

```{r}
pre_imputation_data <- merged_data %>%
  filter(!is.na(rh_strain))
```

```{r}
completed_data <- complete(imputed_data, 1)
merged_data$rh_strain[is.na(merged_data$rh_strain)] <- 
  completed_data$rh_strain[is.na(merged_data$rh_strain)]
```

Sensitivity Analysis to Justify Multiple Imputation of rh_strain: The bar plot comparing observed and imputed values of rh_strain indicates that the proportions of levels (0 vs. 1) are reasonably consistent across both groups. This further suggests that the imputation preserved the original distribution of the data. 

```{r}
merged_data %>%
  mutate(rh_strain_flag = ifelse(is.na(impute_data$rh_strain), "Imputed", "Observed")) %>%
  group_by(rh_strain_flag, rh_strain) %>%
  summarise(n = n(), .groups = "drop") %>%
  group_by(rh_strain_flag) %>%
  mutate(prop = n / sum(n)) %>%
  ggplot(aes(x = factor(rh_strain), y = prop, fill = rh_strain_flag)) +
  geom_bar(stat = "identity", position = "dodge") +
  labs(title = "Proportion of Right Heart Strain (Observed vs Imputed)",
       x = "rh_strain", y = "Proportion") +
  theme_minimal()
```

To assess whether imputation influenced model results, we fit two logistic regression models: one using the complete-case data (before imputation) and another using the full dataset after imputing missing right heart strain values. The results show that the interpretation of the model remains consistent before and after imputation: The rh_strain variable is not statistically significant predictor of pulmonary embolism in either model at a alpha value of 0.10. In addition, the coefficient for rh_strain changed insignificantly after imputation (-1.329 to -1.345) which indicates that the imputation of values did not significantly change the variable's relationship with the outcome variable, pe_status. This validates the appropriateness of the imputation procedure.

```{r}
impute_vars <- c( "rh_strain", "systolic", "map_value", "revised_geneva_score", "well_score") 
logit_formula <- as.formula(paste("pe_status ~", paste(impute_vars, collapse = " + ")))

model_complete <- glm(logit_formula,
                      data = pre_imputation_data,
                      family = binomial)

model_imputed <- glm(logit_formula,
                     data = merged_data,
                     family = binomial)

summary_complete <- summary(model_complete)
summary_imputed <- summary(model_imputed)

logit_table <- data.frame(
  Coef_Complete = round(summary_complete$coefficients[, "Estimate"], 3),
  P_Complete = round(summary_complete$coefficients[, "Pr(>|z|)"], 3),
  Coef_Imputed = round(summary_imputed$coefficients[, "Estimate"], 3),
  P_Imputed = round(summary_imputed$coefficients[, "Pr(>|z|)"], 3)
)


logit_table_sorted <- logit_table[order(logit_table$P_Imputed), ]

logit_table_sorted %>%
  kable("html", caption = "Logistic Regression Comparison Between Pre and Post Imputated Data") %>%
  kable_styling(bootstrap_options = c("striped", "hover", "condensed"), full_width = FALSE)
```


**Variable: pulmonary_hypertension ** 

The variable "pulmonary_hypertension" has 39.617 percent of its data missing. To assess the nature of this missingness, we examine whether it is associated with the outcome variable, pe_status. A cross-tabulation shows that among individuals diagnosed with PE, 98 had observed rh_strain values while 89 had missing values. Among those without PE, 117 had observed values and 53 had missing values. The significant difference in proportions of missingness in both groups suggest that the absence of data is systematically linked to PE status, supporting the assumption that the data may not be Missing Completely At Random (MCAR). We also examined the PE rate within each group. Among those with observed pulmonary hypertension values, the PE rate is 54.4 percent, whereas it is 37.32 percent among those with missing pulmonary hypertension values. 

```{r}
pulmonary_hypertension_pe_table <- table(is.na(merged_data$pulmonary_hypertension), merged_data$pe_status)
pulmonary_hypertension_pe_df <- as.data.frame.matrix(pulmonary_hypertension_pe_table)
rownames(pulmonary_hypertension_pe_df) <- c("Observed (Not Missing)", "Missing")
pulmonary_hypertension_pe_df <- tibble::rownames_to_column(pulmonary_hypertension_pe_df, var = "Pulmonary Hypertension Missingness")
colnames(pulmonary_hypertension_pe_df)[2:3] <- c("PE Absent (0)", "PE Present (1)")

pulmonary_hypertension_pe_df %>%
  kable("html", caption = "Cross-tabulation of Pulmonary Hypertension Strain Missingness by PE Status") %>%
  kable_styling(bootstrap_options = c("striped", "hover", "condensed"), full_width = FALSE)
```

```{r}
merged_data %>%
  mutate(pe_status_num = as.numeric(as.character(pe_status)),
         pulmonary_hypertension_missing = ifelse(is.na(pulmonary_hypertension), "Missing", "Not Missing")) %>%
  group_by(pulmonary_hypertension_missing) %>%
  summarise(pe_rate = mean(pe_status_num, na.rm = TRUE),
            n = n()) %>%
  kable("html", caption = "Pulmonary Embolism Rate by Pulmonary Hypertension") %>%
  kable_styling(full_width = FALSE, bootstrap_options = c("striped", "hover"))
```


These findings suggest that the missingness in "pulmonary_hypertension" is systematically related to observed data, and thus can be considered MAR. Multiple imputation will conducted rather than complete case analysis. A complete case approach would result in the exclusion of 142 observations, which could reduce statistical power and introduce bias, especially if the missingness is related to the outcome variable, PE status. In order to handle the missing values in SB, we applied Multiple Imputation by Chained Equations (MICE) using logistic regression imputation.

The imputation model included the outcome variable pe_status and additional covariates to preserve the variability and account for potential relationships between pulmonary_hypertension and other predictors. To identify relevant covariates for imputation, we trained a Random Forest classifier on the observed data and ranked the predictors based on their Mean Decrease in Accuracy. From variables with no missingness, we selected top variables which were chosen on predictive strength. 

```{r}
no_missing_vars <- names(merged_data)[colSums(is.na(merged_data)) == 0]
merged_data_no_missing <- merged_data[, no_missing_vars]
```

```{r}
formula_str <- paste(
  "pulmonary_hypertension ~", 
  paste(no_missing_vars[no_missing_vars != "pulmonary_hypertension"], collapse = " + ")
)
rf_formula <- as.formula(formula_str)

rf_data <- merged_data %>% filter(!is.na(pulmonary_hypertension))

set.seed(123)
rf_model <- randomForest(
  formula = rf_formula,
  data = rf_data,
  importance = TRUE, 
  ntree = 500
)

var_importance <- importance(rf_model, type = 1) 
top_vars <- rownames(var_importance)[order(-var_importance[, 1])][1:10]

importance_df <- as.data.frame(importance(rf_model))
importance_df$Variable <- rownames(importance_df)

top10_vars <- importance_df %>%
  arrange(desc(MeanDecreaseAccuracy)) %>%
  select(Variable, MeanDecreaseAccuracy) %>%
  slice(1:10)

top10_vars %>%
  kable("html", caption = "Top 10 Predictors of Pulmonary Hypertension (by Mean Decrease Accuracy)") %>%
  kable_styling(bootstrap_options = c("striped", "hover", "condensed"), full_width = FALSE)
```

The same iterative imputation procedure is applied to the variable pulmonary_hypertension to find the optimal set for imputing missing values. The code is essentially the same as iterative imputation performed for previous variables above. Additionally, we included pe_status (the outcome variable). These covariates, together with pulmonary_hypertension were included in the MICE procedure with 5 imputed datasets and 10 iterations, using the first completed dataset to impute the missing values. Code is not included as it is the same as the one from variables above. 

```{r}
impute_vars <- c("pe_status", "pulmonary_hypertension", "systolic", "copd_history", "age") 
impute_data <- merged_data[, impute_vars]
```

```{r, message=FALSE, warning=FALSE, results='hide'}
capture.output({
  imputed_data <- mice(
    impute_data,
    m = 5,
    method = "logreg",
    maxit = 10,
    seed = 123
  )
}, file = NULL)
```

```{r}
pre_imputation_data <- merged_data %>%
  filter(!is.na(pulmonary_hypertension))
```

```{r}
completed_data <- complete(imputed_data, 1)
merged_data$pulmonary_hypertension[is.na(merged_data$pulmonary_hypertension)] <- 
  completed_data$pulmonary_hypertension[is.na(merged_data$pulmonary_hypertension)]
```

Sensitivity Analysis to Justify Multiple Imputation of pulmonary_hypertension: The bar plot comparing observed and imputed values of rh_strain indicates that the proportions of levels (0 vs. 1) are reasonably consistent across both groups. This further suggests that the imputation preserved the original distribution of the data. 

```{r}
merged_data %>%
  mutate(pulmonary_hypertension_flag = ifelse(is.na(impute_data$pulmonary_hypertension), "Imputed", "Observed")) %>%
  group_by(pulmonary_hypertension_flag, pulmonary_hypertension) %>%
  summarise(n = n(), .groups = "drop") %>%
  group_by(pulmonary_hypertension_flag) %>%
  mutate(prop = n / sum(n)) %>%
  ggplot(aes(x = factor(pulmonary_hypertension), y = prop, fill = pulmonary_hypertension_flag)) +
  geom_bar(stat = "identity", position = "dodge") +
  labs(title = "Proportion of Pulmonary Hypertension (Observed vs Imputed)",
       x = "pulmonary_hypertension", y = "Proportion") +
  theme_minimal()
```

To assess whether imputation influenced model results, we fit two logistic regression models: one using the complete-case data (before imputation) and another using the full dataset after imputing missing pulmonary hypertension values. The results show that the interpretation of the model remains consistent before and after imputation: The pulmonary_hypertension variable is  statistically significant predictor of pulmonary embolism in either model at a alpha value of 0.05. In addition, the coefficient for pulmonary_hypertension changed insignificantly after imputation (-0.780 to -0.633) which indicates that the imputation of values did not significantly change the variable's relationship with the outcome variable, pe_status. This validates the appropriateness of the imputation procedure.

```{r}
impute_vars <- c( "pulmonary_hypertension", "systolic", "copd_history", "age") 
logit_formula <- as.formula(paste("pe_status ~", paste(impute_vars, collapse = " + ")))

model_complete <- glm(logit_formula,
                      data = pre_imputation_data,
                      family = binomial)

model_imputed <- glm(logit_formula,
                     data = merged_data,
                     family = binomial)

summary_complete <- summary(model_complete)
summary_imputed <- summary(model_imputed)

logit_table <- data.frame(
  Coef_Complete = round(summary_complete$coefficients[, "Estimate"], 3),
  P_Complete = round(summary_complete$coefficients[, "Pr(>|z|)"], 3),
  Coef_Imputed = round(summary_imputed$coefficients[, "Estimate"], 3),
  P_Imputed = round(summary_imputed$coefficients[, "Pr(>|z|)"], 3)
)


logit_table_sorted <- logit_table[order(logit_table$P_Imputed), ]

logit_table_sorted %>%
  kable("html", caption = "Logistic Regression Comparison Between Pre and Post Imputated Data") %>%
  kable_styling(bootstrap_options = c("striped", "hover", "condensed"), full_width = FALSE)
```


**Variable: rh_thrombus** 

The variable "rh_thrombus" or right heart thrombus has 39.89 percent of its data missing. To assess the nature of this missingness, we examine whether it is associated with the outcome variable, pe_status. A cross-tabulation shows that among individuals diagnosed with PE, 98 had observed rh_thrombus values while 89 had missing values. Among those without PE, 116 had observed values and 54 had missing values. The significant difference in proportions of missingness in both groups suggest that the absence of data is systematically linked to PE status, supporting the assumption that the data may not be Missing Completely At Random (MCAR). We also examined the PE rate within each group. Among those with observed rh_thrombus values, the PE rate is 54.2 percent, whereas it is 37.76 percent among those with missing rh_thrombus values. 

```{r}
rh_thrombus_pe_table <- table(is.na(merged_data$rh_thrombus), merged_data$pe_status)
rh_thrombus_pe_df <- as.data.frame.matrix(rh_thrombus_pe_table)
rownames(rh_thrombus_pe_df) <- c("Observed (Not Missing)", "Missing")
rh_thrombus_pe_df <- tibble::rownames_to_column(rh_thrombus_pe_df, var = "Right Heart Thrombus Missingness")
colnames(rh_thrombus_pe_df)[2:3] <- c("PE Absent (0)", "PE Present (1)")

rh_thrombus_pe_df %>%
  kable("html", caption = "Cross-tabulation of Right Hand Thrombus Missingness by PE Status") %>%
  kable_styling(bootstrap_options = c("striped", "hover", "condensed"), full_width = FALSE)
```

```{r}
merged_data %>%
  mutate(pe_status_num = as.numeric(as.character(pe_status)),
         rh_thrombus_missing = ifelse(is.na(rh_thrombus), "Missing", "Not Missing")) %>%
  group_by(rh_thrombus_missing) %>%
  summarise(pe_rate = mean(pe_status_num, na.rm = TRUE),
            n = n()) %>%
  kable("html", caption = "Right Hand Thrombus Rate by Pulmonary Hypertension") %>%
  kable_styling(full_width = FALSE, bootstrap_options = c("striped", "hover"))
```


These findings suggest that the missingness in "rh_thrombus" is systematically related to observed data, and thus can be considered MAR. Multiple imputation will conducted rather than complete case analysis. A complete case approach would result in the exclusion of 143 observations, which could reduce statistical power and introduce bias, especially if the missingness is related to the outcome variable, PE status. In order to handle the missing values in this variable, we applied Multiple Imputation by Chained Equations (MICE) using logistic regression imputation.

The imputation model included the outcome variable pe_status and additional covariates to preserve the variability and account for potential relationships between rh_thrombus and other predictors. To identify relevant covariates for imputation, we trained a Random Forest classifier on the observed data and ranked the predictors based on their Mean Decrease in Accuracy. From variables with no missingness, we selected top variables which were chosen on predictive strength. 

```{r}
no_missing_vars <- names(merged_data)[colSums(is.na(merged_data)) == 0]
merged_data_no_missing <- merged_data[, no_missing_vars]
```

```{r}
formula_str <- paste(
  "rh_thrombus ~", 
  paste(no_missing_vars[no_missing_vars != "rh_thrombus"], collapse = " + ")
)
rf_formula <- as.formula(formula_str)

rf_data <- merged_data %>% filter(!is.na(rh_thrombus))

set.seed(123)
rf_model <- randomForest(
  formula = rf_formula,
  data = rf_data,
  importance = TRUE, 
  ntree = 500
)

var_importance <- importance(rf_model, type = 1) 
top_vars <- rownames(var_importance)[order(-var_importance[, 1])][1:10]

importance_df <- as.data.frame(importance(rf_model))
importance_df$Variable <- rownames(importance_df)

top10_vars <- importance_df %>%
  arrange(desc(MeanDecreaseAccuracy)) %>%
  select(Variable, MeanDecreaseAccuracy) %>%
  slice(1:10)

top10_vars %>%
  kable("html", caption = "Top 10 Predictors of Right Hand Thrombus (by Mean Decrease Accuracy)") %>%
  kable_styling(bootstrap_options = c("striped", "hover", "condensed"), full_width = FALSE)
```

The same iterative imputation procedure is applied to the variable rh_thrombus to find the optimal set for imputing missing values. The code is essentially the same as iterative imputation performed for previous variables above. Additionally, we included pe_status (the outcome variable). These covariates, together with rh_thrombus were included in the MICE procedure with 5 imputed datasets and 10 iterations, using the first completed dataset to impute the missing values. Code is not included as it is the same as the one from variables above. 

```{r}
impute_vars <- c("pe_status", "rh_thrombus", "dvt_pe", "revised_geneva_score", "elevated_bnp") 
impute_data <- merged_data[, impute_vars]
```

```{r, message=FALSE, warning=FALSE, results='hide'}
capture.output({
  imputed_data <- mice(
    impute_data,
    m = 5,
    method = "logreg",
    maxit = 10,
    seed = 123
  )
}, file = NULL)
```

```{r}
pre_imputation_data <- merged_data %>%
  filter(!is.na(rh_thrombus))
```

```{r}
completed_data <- complete(imputed_data, 1)
merged_data$rh_thrombus[is.na(merged_data$rh_thrombus)] <- 
  completed_data$rh_thrombus[is.na(merged_data$rh_thrombus)]
```

Sensitivity Analysis to Justify Multiple Imputation of rh_thrombus: The bar plot comparing observed and imputed values of rh_strain indicates that the proportions of levels (0 vs. 1) are reasonably consistent across both groups. This further suggests that the imputation preserved the original distribution of the data. 

```{r}
merged_data %>%
  mutate(rh_thrombus_flag = ifelse(is.na(impute_data$rh_thrombus), "Imputed", "Observed")) %>%
  group_by(rh_thrombus_flag, rh_thrombus) %>%
  summarise(n = n(), .groups = "drop") %>%
  group_by(rh_thrombus_flag) %>%
  mutate(prop = n / sum(n)) %>%
  ggplot(aes(x = factor(rh_thrombus), y = prop, fill = rh_thrombus_flag)) +
  geom_bar(stat = "identity", position = "dodge") +
  labs(title = "Proportion of Right Hand Thrombus (Observed vs Imputed)",
       x = "rh_thrombus", y = "Proportion") +
  theme_minimal()
```

To assess whether imputation influenced model results, we fit two logistic regression models: one using the complete-case data (before imputation) and another using the full dataset after imputing missing rh_thrombus values. The results show that the interpretation of the model remains consistent before and after imputation: The rh_thrombus variable is not statistically significant predictor of pulmonary embolism in either model at a alpha value of 0.10. In addition, the coefficient for rh_thrombus changed insignificantly after imputation (-14.652 to -15.281) which indicates that the imputation of values did not significantly change the variable's relationship with the outcome variable, pe_status. This validates the appropriateness of the imputation procedure.

```{r}
impute_vars <- c("rh_thrombus", "dvt_pe", "revised_geneva_score", "elevated_bnp") 
logit_formula <- as.formula(paste("pe_status ~", paste(impute_vars, collapse = " + ")))

model_complete <- glm(logit_formula,
                      data = pre_imputation_data,
                      family = binomial)

model_imputed <- glm(logit_formula,
                     data = merged_data,
                     family = binomial)

summary_complete <- summary(model_complete)
summary_imputed <- summary(model_imputed)

logit_table <- data.frame(
  Coef_Complete = round(summary_complete$coefficients[, "Estimate"], 3),
  P_Complete = round(summary_complete$coefficients[, "Pr(>|z|)"], 3),
  Coef_Imputed = round(summary_imputed$coefficients[, "Estimate"], 3),
  P_Imputed = round(summary_imputed$coefficients[, "Pr(>|z|)"], 3)
)


logit_table_sorted <- logit_table[order(logit_table$P_Imputed), ]

logit_table_sorted %>%
  kable("html", caption = "Logistic Regression Comparison Between Pre and Post Imputated Data") %>%
  kable_styling(bootstrap_options = c("striped", "hover", "condensed"), full_width = FALSE)
```




**Variable: rv_hypokines_on_echo** 

The variable "rv_hypokines_on_echo" has 39.89 percent of its data missing. To assess the nature of this missingness, we examine whether it is associated with the outcome variable, pe_status. A cross-tabulation shows that among individuals diagnosed with PE, 97 had observed rv_hypokines_on_echo values while 90 had missing values. Among those without PE, 117 had observed values and 53 had missing values. The significant difference in proportions of missingness in both groups suggest that the absence of data is systematically linked to PE status, supporting the assumption that the data may not be Missing Completely At Random (MCAR). We also examined the PE rate within each group. Among those with observed rv_hypokines_on_echo values, the PE rate is 54.67 percent, whereas it is 37.06 percent among those with missing rv_hypokines_on_echo values. 

```{r}
rv_hypokines_on_echo_pe_table <- table(is.na(merged_data$rv_hypokines_on_echo), merged_data$pe_status)
rv_hypokines_on_echo_pe_df <- as.data.frame.matrix(rv_hypokines_on_echo_pe_table)
rownames(rv_hypokines_on_echo_pe_df) <- c("Observed (Not Missing)", "Missing")
rv_hypokines_on_echo_pe_df <- tibble::rownames_to_column(rv_hypokines_on_echo_pe_df, var = "Right Ventricular Hypokinesis Missingness")
colnames(rv_hypokines_on_echo_pe_df)[2:3] <- c("PE Absent (0)", "PE Present (1)")

rv_hypokines_on_echo_pe_df %>%
  kable("html", caption = "Cross-tabulation of Right Ventricular Hypokinesis Missingness by PE Status") %>%
  kable_styling(bootstrap_options = c("striped", "hover", "condensed"), full_width = FALSE)
```


```{r}
merged_data %>%
  mutate(pe_status_num = as.numeric(as.character(pe_status)),
         rv_hypokines_on_echo_missing = ifelse(is.na(rv_hypokines_on_echo), "Missing", "Not Missing")) %>%
  group_by(rv_hypokines_on_echo_missing) %>%
  summarise(pe_rate = mean(pe_status_num, na.rm = TRUE),
            n = n()) %>%
  kable("html", caption = "Pulmonary Embolism Rate by Right Ventricular Hypokinesis") %>%
  kable_styling(full_width = FALSE, bootstrap_options = c("striped", "hover"))
```


These findings suggest that the missingness in "rv_hypokines_on_echo" is systematically related to observed data, and thus can be considered MAR. Multiple imputation will conducted rather than complete case analysis. A complete case approach would result in the exclusion of 143 observations, which could reduce statistical power and introduce bias, especially if the missingness is related to the outcome variable, PE status. In order to handle the missing values in this variable, we applied Multiple Imputation by Chained Equations (MICE) using logistic regression imputation.

The imputation model included the outcome variable pe_status and additional covariates to preserve the variability and account for potential relationships between rv_hypokines_on_echo and other predictors. To identify relevant covariates for imputation, we trained a Random Forest classifier on the observed data and ranked the predictors based on their Mean Decrease in Accuracy. From variables with no missingness, we selected top variables which were chosen on predictive strength. 

```{r}
no_missing_vars <- names(merged_data)[colSums(is.na(merged_data)) == 0]
merged_data_no_missing <- merged_data[, no_missing_vars]
```

```{r}
formula_str <- paste(
  "rv_hypokines_on_echo ~", 
  paste(no_missing_vars[no_missing_vars != "rv_hypokines_on_echo"], collapse = " + ")
)
rf_formula <- as.formula(formula_str)

rf_data <- merged_data %>% filter(!is.na(rv_hypokines_on_echo))

set.seed(123)
rf_model <- randomForest(
  formula = rf_formula,
  data = rf_data,
  importance = TRUE, 
  ntree = 500
)

var_importance <- importance(rf_model, type = 1) 
top_vars <- rownames(var_importance)[order(-var_importance[, 1])][1:10]

importance_df <- as.data.frame(importance(rf_model))
importance_df$Variable <- rownames(importance_df)

top10_vars <- importance_df %>%
  arrange(desc(MeanDecreaseAccuracy)) %>%
  select(Variable, MeanDecreaseAccuracy) %>%
  slice(1:10)

top10_vars %>%
  kable("html", caption = "Top 10 Predictors of Right Ventricular Hypokinesis (by Mean Decrease Accuracy)") %>%
  kable_styling(bootstrap_options = c("striped", "hover", "condensed"), full_width = FALSE)
```

The same iterative imputation procedure is applied to the variable rv_hypokines_on_echo to find the optimal set for imputing missing values. The code is essentially the same as iterative imputation performed for previous variables above. Additionally, we included pe_status (the outcome variable). These covariates, together with rv_hypokines_on_echo were included in the MICE procedure with 5 imputed datasets and 10 iterations, using the first completed dataset to impute the missing values. Code is not included as it is the same as the one from variables above. 


```{r}
impute_vars <- c("pe_status", "rv_hypokines_on_echo", "map_value", "rh_strain", "elevated_bnp")
impute_data <- merged_data[, impute_vars]
```

```{r, message=FALSE, warning=FALSE, results='hide'}
capture.output({
  imputed_data <- mice(
    impute_data,
    m = 5,
    method = "logreg",
    maxit = 10,
    seed = 123
  )
}, file = NULL)
```

```{r}
pre_imputation_data <- merged_data %>%
  filter(!is.na(rv_hypokines_on_echo))
```

```{r}
completed_data <- complete(imputed_data, 1)
merged_data$rv_hypokines_on_echo[is.na(merged_data$rv_hypokines_on_echo)] <- 
  completed_data$rv_hypokines_on_echo[is.na(merged_data$rv_hypokines_on_echo)]
```

Sensitivity Analysis to Justify Multiple Imputation of rv_hypokines_on_echo: The bar plot comparing observed and imputed values of rv_hypokines_on_echo indicates that the proportions of levels (0 vs. 1) are reasonably consistent across both groups. This further suggests that the imputation preserved the original distribution of the data. 

```{r}
merged_data %>%
  mutate(rv_hypokines_on_echo_flag = ifelse(is.na(impute_data$rv_hypokines_on_echo), "Imputed", "Observed")) %>%
  group_by(rv_hypokines_on_echo_flag, rv_hypokines_on_echo) %>%
  summarise(n = n(), .groups = "drop") %>%
  group_by(rv_hypokines_on_echo_flag) %>%
  mutate(prop = n / sum(n)) %>%
  ggplot(aes(x = factor(rv_hypokines_on_echo), y = prop, fill = rv_hypokines_on_echo_flag)) +
  geom_bar(stat = "identity", position = "dodge") +
  labs(title = "Proportion ofRight Ventricular Hypokinesis (Observed vs Imputed)",
       x = "rh_thrombus", y = "Proportion") +
  theme_minimal()
```

To assess whether imputation influenced model results, we fit two logistic regression models: one using the complete-case data (before imputation) and another using the full dataset after imputing missing rv_hypokines_on_echo values. The results show that the interpretation of the model remains consistent before and after imputation: The rv_hypokines_on_echo variable is statistically significant predictor of pulmonary embolism in either model at a alpha value of 0.05. In addition, the coefficient for rv_hypokines_on_echo changed insignificantly after imputation (-1.185 to -0.986) which indicates that the imputation of values did not significantly change the variable's relationship with the outcome variable, pe_status. This validates the appropriateness of the imputation procedure.

```{r}
impute_vars <- c( "rv_hypokines_on_echo", "map_value", "rh_strain", "elevated_bnp")
logit_formula <- as.formula(paste("pe_status ~", paste(impute_vars, collapse = " + ")))

model_complete <- glm(logit_formula,
                      data = pre_imputation_data,
                      family = binomial)

model_imputed <- glm(logit_formula,
                     data = merged_data,
                     family = binomial)

summary_complete <- summary(model_complete)
summary_imputed <- summary(model_imputed)

logit_table <- data.frame(
  Coef_Complete = round(summary_complete$coefficients[, "Estimate"], 3),
  P_Complete = round(summary_complete$coefficients[, "Pr(>|z|)"], 3),
  Coef_Imputed = round(summary_imputed$coefficients[, "Estimate"], 3),
  P_Imputed = round(summary_imputed$coefficients[, "Pr(>|z|)"], 3)
)


logit_table_sorted <- logit_table[order(logit_table$P_Imputed), ]

logit_table_sorted %>%
  kable("html", caption = "Logistic Regression Comparison Between Pre and Post Imputated Data") %>%
  kable_styling(bootstrap_options = c("striped", "hover", "condensed"), full_width = FALSE)
```


**Variables: echo_rvsp_value and rvsp_based_age ** 

Based on the cross-tabulation output, we observe that among individuals diagnosed with pulmonary embolism (PE), 82 had observed values for echo_rvsp_value (Echocardiographic RVSP Right Ventricular Systoic Pressure data), while 88 had missing entries (48 percent missing). In contrast, among those not diagnosed with PE, only 57 had observed values, and 130 had missing values (70 percent missing). This indicates that missingness in echo_rvsp_value is more prevalent among individuals without PE. The unequal distribution of missing values between the PE and non-PE groups suggests that the missingness is not completely at random (not MCAR), and may be associated with PE status. 

```{r}
echo_pe_table <- table(is.na(merged_data$echo_rvsp_value), merged_data$pe_status)
echo_pe_df <- as.data.frame.matrix(echo_pe_table)
rownames(echo_pe_df) <- c("Observed (Not Missing)", "Missing")
echo_pe_df <- tibble::rownames_to_column(echo_pe_df, var = "Echo RVSP Missingness")
colnames(echo_pe_df)[2:3] <- c("PE Absent (0)", "PE Present (1)")
echo_pe_df %>%
  kable("html", caption = "Cross-tabulation of Echo RVSP Missingness by PE Status") %>%
  kable_styling(bootstrap_options = c("striped", "hover", "condensed"), full_width = FALSE)
```


Furthermore, the chi-square test of independence yielded a small p-value (0.0008771), indicating a significant association between missingness in echo_rvsp_value and the outcome variable, pe_status. This further supports the conclusion that missingness in this variable is not completely at random (not MCAR). The Little's MCAR test also showcases a small p-value, confirming that missingess is not MCAR and that the chi-square test results are valid. 

```{r}
chisq.test(table(is.na(merged_data$echo_rvsp_value), merged_data$pe_status))
mcar_test(merged_data[,c("echo_rvsp_value","pe_status")]) 
```

To investigate the missingness mechanism of the echo_rvsp_value variable, we conducted a logistic regression analysis to determine whether the probability of missingness is associated with other observed variables. We first created a binary indicator variable, missing_echo_rvsp_value, where 1 indicates a missing value and 0 indicates an observed one. To avoid introducing bias or excessive row deletion (the glm() function automatically excludes rows with any missing values), we restricted our analysis to a subset of variables that have no missing data. The resulting logistic regression model revealed that the variable pe_status was a significant predictor of missingness, suggesting that the missingness in echo_rvsp_value is not completely at random (MCAR) and is more likely Missing at Random (MAR), as it appears to depend on observed data. Therefore, we proceed with multiple imputation rather than complete case analysis to retain statistical power and reduce bias.

```{r}
merged_data$missing_echo_rvsp_value <- as.numeric(is.na(merged_data$echo_rvsp_value))
```

```{r}
no_missing_vars <- names(merged_data)[colSums(is.na(merged_data)) == 0]
merged_data_no_missing <- merged_data[, no_missing_vars]
```

```{r}
excluded_vars <- c("echo_rvsp_value", "missing_echo_rvsp_value", "rvsp_based_age")
predictor_vars <- setdiff(names(merged_data_no_missing), excluded_vars)

formula_str <- paste("missing_echo_rvsp_value ~", paste(predictor_vars, collapse = " + "))
formula <- as.formula(formula_str)

missingness_model <- glm(formula, data = merged_data_no_missing, family = binomial)

significant_vars <- tidy(missingness_model) %>%
  filter(term != "(Intercept)", p.value < 0.1) %>%
  arrange(p.value)

significant_vars %>%
  mutate(p.value = signif(p.value, 3)) %>%
  kable("html", caption = "Significant Predictors of Missingness in echo_rvsp_value") %>%
  kable_styling(bootstrap_options = c("striped", "hover", "condensed"), full_width = FALSE)
```


Imputation for echo_rvsp_value: Missing values in this variable were imputed using Predictive Mean Matching (PMM) via the MICE algorithm. The predictor variables used in the imputation include both: (1) variables most strongly associated with missingness in echo_rvsp_value with a p-value of less than 0.10 (as detailed in the table above), and (2) clinically relevant predictors known to be associated with both the outcome variable pe_status and with RVSP (Right Ventricular Systolic Pressure) which is a measurement that estimates the pressure in the right ventricle of the heart. All these variables have no missing values. 

To address the skewed distribution of RVSP, values for echo_rvsp_value were log-transformed using log1p() prior to imputation. PMM was then applied over 20 imputed datasets and 50 iterations. After imputation, the values were back-transformed to their original scale using expm1(). Finally, the binary variable rvsp_based_age was derived from the imputed echo_rvsp_value, coded as 1 for values greater than 40 mmHg, and 0 otherwise.

```{r}
mandatory_predictors <- c(
  "acute_resp", "haemoptysis", "immobilization_surgery", "pe_status", "admission_year"
)
clinical_predictors <- c(
  "heart_rate", "diastolic", "systolic")
  
  impute_predictors <- c(mandatory_predictors, clinical_predictors)
```

```{r}
merged_data$log_rvsp <- log1p(merged_data$echo_rvsp_value)
impute_data <- merged_data[, c("log_rvsp", impute_predictors)]
```

```{r}
pred_matrix <- make.predictorMatrix(impute_data)
pred_matrix[,] <- 0  
pred_matrix["log_rvsp", impute_predictors] <- 1
```

```{r, message=FALSE, warning=FALSE, results='hide'}
capture.output(
  imp <- mice(impute_data,
              method = "pmm", 
              predictorMatrix = pred_matrix,
              m = 20, 
              maxit = 50,
              seed = 123),
  file = NULL
)
```

```{r}
log_rvsp<- complete(imp, 1)$log_rvsp
merged_data$echo_rvsp_value <- expm1(log_rvsp)

merged_data$echo_rvsp_value <- as.numeric(merged_data$echo_rvsp_value)
merged_data$rvsp_based_age <- factor(
  ifelse(merged_data$echo_rvsp_value > 40, 1, 0),
  levels = c(0, 1)
)
```

Validating Imputation of echo_rvsp_value and rvsp_based_age: The summary table presents descriptive statistics for echo_rvsp_value by imputed_flag. The observed and imputed groups show similar central tendencies, with mean values of 39.87 (observed) and 38.52 (imputed), and median values of 36. Additionally, results from the Welch two-sample t-test and Wilcoxon rank-sum test yield p-values greater than 0.05, indicating no statistically significant difference in the means or distributions of echo_rvsp-value between the observed and imputed groups. This conclusion is further supported by the density plot, which shows substantial overlap between the two distributions and no notable shift post-imputation. Taken together, these findings confirm that the imputation of echo_rvsp_value was successful and did not introduce meaningful bias.

```{r}
imputed_df <- complete(imp, 1)
imputed_df$imputed_flag <- ifelse(is.na(merged_data$log_rvsp), "Imputed", "Observed")

imputed_df$echo_rvsp_value<- expm1(imputed_df$log_rvsp)

summary_table <- imputed_df %>%
  mutate(imputed_flag = factor(imputed_flag, levels = c("Observed", "Imputed"))) %>%
  group_by(imputed_flag) %>%
  summarise(
    Mean = round(mean(echo_rvsp_value, na.rm = TRUE), 2),
    Median = round(median(echo_rvsp_value, na.rm = TRUE), 2),
    SD = round(sd(echo_rvsp_value, na.rm = TRUE), 2),
    N = n(),
    .groups = 'drop'
  )

summary_table %>%
  kable("html", caption = "Summary of RVSP Levels by Imputation Status") %>%
  kable_styling(bootstrap_options = c("striped", "hover", "condensed", "responsive"))
```


```{r}
t.test(echo_rvsp_value ~ imputed_flag, data = imputed_df)
```


```{r}
wilcox.test(echo_rvsp_value ~ imputed_flag, data = imputed_df)
```

```{r}
ggplot(imputed_df, aes(x = echo_rvsp_value, fill = imputed_flag)) +
  geom_density(alpha = 0.4) +
  labs(
    title = "Distribution of Observed vs. Imputed RVSP Levels",
    x = "RVSP Levels",
    y = "Density",
    fill = "Value Type"
  ) +
  theme_minimal()

merged_data <- merged_data %>% select(-log_rvsp)
merged_data <- merged_data %>% select(-missing_echo_rvsp_value) 
```


**Finalized Data Set ** 

After dealing with missing values present in various variables in the merged_data dataset, the finalized data set that will be used for analysis is as follows. There are 357 total observations (rows) and 47 columns (variables) in the final dataset. 

```{r}
head(merged_data) 
str(merged_data) 
dim(merged_data)
```

# Part C: Descriptive Statistics (Univariate): Identifying Potential Outliers/Uncaptured Missing Values (No Stratification) 

**1. Categorical Variables ** 

Statistical Summaries (Counts, Frequencies, and Percentages) 

Visualizations: Bar Graphs, Pie Charts 

```{r}

```


**2. Continuous Variables (Numerical):** 

Statistical Summaries 

Visaulizations 

Identifying Outliers 


#Part D: Stratified Analysis (Bivariate): Stratify individuals by the presence of Pulmonary Embolism 




#Part E: Multivariate Analysis (Logistic Regression): First Research Question 

First Research Question: 





