

```{r}
library(readxl)
library(dplyr)
library(readr)
library(tibble) 
library(ggplot2) 
library(knitr) 
library(kableExtra) 
library(tidyr)
library(mice)
library(naniar)
library(randomForest)
```

# Part A: Data Wrangling
**Construction of the merged_data Dataset ** 

```{r}
full_data <- "~/Documents/Desert Valley Research /desert_valley.xlsx"
cases <- read_excel(full_data, sheet = 1)
controls <- read_excel(full_data, sheet = 2)
```


**New Variables: ** Three new variables were derived to enhance interpretability: systolic, diastolic, and bmi_marker. The systolic and diastolic variables were extracted by decomposing the original blood_pressure variable, which contained combined systolic/diastolic readings. Additionally, bmi_marker was created by categorizing the continuous bmi variable into six clinically meaningful BMI categories. 

```{r}
cases$heart_rate_geneva <- ifelse(
  cases$heart_rate >= 95, 3,
  ifelse(cases$heart_rate >= 75 & cases$heart_rate <= 94, 2, 1)
)

controls$heart_rate_geneva <- ifelse(
  controls$heart_rate >= 95, 3,
  ifelse(controls$heart_rate >= 75 & controls$heart_rate <= 94, 2, 1)
) 

bp_split_cases <- strsplit(cases$blood_pressure, "/")
cases$systolic <- as.numeric(sapply(bp_split_cases, `[`, 1))
cases$diastolic <- as.numeric(sapply(bp_split_cases, `[`, 2))

bp_split_controls <- strsplit(controls$blood_pressure, "/")
controls$systolic <- as.numeric(sapply(bp_split_controls, `[`, 1))
controls$diastolic <- as.numeric(sapply(bp_split_controls, `[`, 2))

cases$blood_pressure <- NULL
controls$blood_pressure <- NULL
```

```{r}
cases$bmi_marker <- cut(
  cases$bmi,
  breaks = c(-Inf, 18.5, 24.9, 29.9, 34.9, 39.9, Inf),
  labels = c(
    "Underweight", 
    "Normal", 
    "Overweight", 
    "Obesity Class 1", 
    "Obesity Class 2", 
    "Obesity Class 3"
  ),
  right = TRUE
)

controls$bmi_marker <- cut(
  controls$bmi,
  breaks = c(-Inf, 18.5, 24.9, 29.9, 34.9, 39.9, Inf),
  labels = c(
    "Underweight", 
    "Normal", 
    "Overweight", 
    "Obesity Class 1", 
    "Obesity Class 2", 
    "Obesity Class 3"
  ),
  right = TRUE
)
```

```{r}
cases$age_group <- cut(
  cases$age,
  breaks = c(18, 30, 40, 50, 60, 70, 80, 100),
  labels = c("19–30", "31–40", "41–50", "51–60", "61–70", "71–80", "81+"),
  right = TRUE,
  include.lowest = TRUE
)

controls$age_group <- cut(
  controls$age,
  breaks = c(18, 30, 40, 50, 60, 70, 80, 100),
  labels = c("19–30", "31–40", "41–50", "51–60", "61–70", "71–80", "81+"),
  right = TRUE,
  include.lowest = TRUE
)
```

**Handling Labeling for Binary Variables: ** For all binary (two-level factor) variables in the original dataset, data entries were standardized for consistency. Values coded as “3” were recoded as NA to indicate missingness, while values coded as “2” were recoded as 0. As a result, all binary variables follow a unified labeling convention: 1 = Yes, 0 = No, and NA = Missing.

```{r}
clean_binary_variable <- function(df, varname) {
  df[[varname]][df[[varname]] == 3] <- NA
  df[[varname]] <- factor(ifelse(df[[varname]] == 2, 0, df[[varname]]))
  return(df)
}

binary_vars <- c(
  "gender", "smoking_history", "copd_history", "emphysema_ct", "copd_exacerbation",
  "positive_d_dimer", "covid", "elevated_bnp", "elevated_troponin",
  "dvt_in_lower_extremity", "rvsp_based_age", "rv_hypokines_on_echo",
  "pulmonary_hypertension", "rh_thrombus", "rh_strain", "rv_lv_abnormality",
  "tricuspid_regurgitation", "rv_edd_abnormality", "septal_bowing", 
  "clinical_symptoms_dvt", "pe_more_likely", "heart_rate_100", 
  "immobilization_surgery", "dvt_pe", "haemoptysis", "malignancy",   
  "surgery_fracture", "active_malignancy", "unilateral_lower_limb_pain", 
  "pain_lower_limb_edema", "acute_resp"
)

for (var in binary_vars) {
  cases <- clean_binary_variable(cases, var)
}

for (var in binary_vars) {
  controls <- clean_binary_variable(controls, var)
}
```


**Handling Numerical Variables with Limited or Censored Data: ** Variables such as d_dimer_value, bnp_levels, troponin_levels, troponin_levels_hs, and map_value contained values reported with inequality signs (ex: “<0.015”). To standardize these entries for quantitative analysis, we parsed and retained only the numeric portion—interpreting values like “<0.015” as 0.015. 

```{r}
numeric_vars <- c("d_dimer_value", "bnp_levels", "troponin_levels", "troponin_levels_hs") 

cases[numeric_vars] <- lapply(cases[numeric_vars], parse_number)
controls[numeric_vars] <- lapply(controls[numeric_vars], parse_number)
```

```{r}
cases$map_value <- parse_number(trimws(cases$map_value))
controls$map_value <- parse_number(trimws(controls$map_value))
```

**Variable Type Conversion:** Each variable was converted to its appropriate data type to ensure accurate analysis. Continuous variables were retained or converted as numeric types, while categorical variables, particularly binary indicators, were converted into factors with consistent labeling (0 = No, 1 = Yes). 

```{r}
convert_variables <- function(df) {
  df <- df %>%
    mutate(
      pe_status = factor(pe_status), 
      age = as.numeric(age),
      gender = factor(gender), 
      race = factor(race), 
      bmi = as.numeric(bmi),
      bmi_marker = as.factor(bmi_marker), 

      smoking_history = factor(smoking_history),
      copd_history = factor(copd_history),
      emphysema_ct = factor(emphysema_ct),
      copd_exacerbation = factor(copd_exacerbation), 
      
      positive_d_dimer = factor(positive_d_dimer),
      d_dimer_value = as.numeric(d_dimer_value),
      
      covid = factor(covid),
      elevated_bnp = factor(elevated_bnp),
      bnp_levels = as.numeric(bnp_levels),
      elevated_troponin = factor(elevated_troponin),
      troponin_levels = as.numeric(troponin_levels),
      troponin_levels_hs = as.numeric(troponin_levels_hs),
      dvt_in_lower_extremity = factor(dvt_in_lower_extremity),
    
      rvsp_based_age = factor(rvsp_based_age),
      echo_rvsp_value = as.numeric(echo_rvsp_value),
      rv_hypokines_on_echo = factor(rv_hypokines_on_echo),
      pulmonary_hypertension = factor(pulmonary_hypertension),
      rh_thrombus = factor(rh_thrombus),
      rh_strain = factor(rh_strain),
      rv_to_lv_ratio = as.numeric(rv_to_lv_ratio),
      rv_lv_abnormality = factor(rv_lv_abnormality),
      
      tricuspid_regurgitation = factor(tricuspid_regurgitation),
      rv_edd_cm = as.numeric(rv_edd_cm),
      rv_edd_abnormality = factor(rv_edd_abnormality),
      septal_bowing = factor(septal_bowing),

      clinical_symptoms_dvt = factor(clinical_symptoms_dvt),
      pe_more_likely = factor(pe_more_likely),
      heart_rate_100 = factor(heart_rate_100),
      heart_rate = as.numeric(heart_rate),
      immobilization_surgery = factor(immobilization_surgery),
      dvt_pe = factor(dvt_pe),
      haemoptysis  = factor(haemoptysis),
      malignancy = factor(malignancy),

      surgery_fracture = factor(surgery_fracture),
      active_malignancy = factor(active_malignancy),
      unilateral_lower_limb_pain = factor(unilateral_lower_limb_pain),
      pain_lower_limb_edema = factor(pain_lower_limb_edema), 
      heart_rate_geneva = factor(heart_rate_geneva),
      
      well_score = as.numeric(well_score), 
      revised_geneva_score = as.numeric(revised_geneva_score), 
      admission_year = factor(admission_year), 
      map_value = as.numeric(map_value), 
      acute_resp = factor(acute_resp),
      diastolic = as.numeric(diastolic),
      systolic = as.numeric(systolic), 
      age_group = as.factor(age_group)
  
    )
  
  return(df)
}

cases <- convert_variables(cases)
controls <- convert_variables(controls)
```


**Merging the Cases and Controls Data: ** The final analytic dataset, named merged_data, comprises 366 observations across 50 variables. This dataset integrates individuals from both the case and control groups: 176 individuals diagnosed with pulmonary embolism (PE) and 190 individuals without PE. The outcome variable, pe_status, indicates case-control status, where a value of 1 denotes a case (PE present) and 0 denotes a control (PE absent).

```{r}
merged_data <- bind_rows(cases, controls) 
merged_data <- merged_data %>%
  select(-c(troponin_levels, troponin_levels_hs))
```

```{r}
str(merged_data)
```

# Part B: Cleaning Data: Dealing with Missing Values 

For the 52 variables presented, let us run a summary to see what percentage of the data are present in each variable: 

```{r}
missing_summary <- merged_data %>%
  summarise(across(everything(), ~mean(is.na(.)) * 100)) %>%
  pivot_longer(everything(), names_to = "Variable", values_to = "Percent_Missing") %>%
  filter(Percent_Missing > 0) %>%
  mutate(Category = case_when(
    Percent_Missing < 5 ~ "<5% Missing",
    Percent_Missing >= 5 & Percent_Missing <= 20 ~ "5–20% Missing",
    Percent_Missing > 20 ~ ">20% Missing"
  )) %>%
  arrange(desc(Percent_Missing))

minimal_missing <- missing_summary %>% filter(Category == "<5% Missing")
moderate_missing <- missing_summary %>% filter(Category == "5–20% Missing")
high_missing <- missing_summary %>% filter(Category == ">20% Missing")


minimal_missing %>%
  select(Variable, Percent_Missing) %>%
  kable("html", caption = "Less than 5% Missing") %>%
  kable_styling(full_width = FALSE, bootstrap_options = c("striped", "hover"))

moderate_missing %>%
  select(Variable, Percent_Missing) %>%
  kable("html", caption = "5–20% Missing") %>%
  kable_styling(full_width = FALSE, bootstrap_options = c("striped", "hover"))

high_missing %>%
  select(Variable, Percent_Missing) %>%
  kable("html", caption = "Greater than 20% Missing") %>%
  kable_styling(full_width = FALSE, bootstrap_options = c("striped", "hover"))
```

**Handling Missingess of Variables with <5% Missing Data ** 

For variables with less than 5% missingness, it is important to determine the underlying mechanism of missing data (MCAR, MAR, or NMAR) in order to select the appropriate method for handling it. Specifically, we assess whether the probability of missingness is independent of both observed and unobserved variables, which is a condition known as Missing Completely at Random (MCAR). To evaluate this, we applied Little’s MCAR test, which statistically compares means and covariances across groups defined by patterns of missing data. This test uses a chi-squared statistic to assess whether there are significant differences between groups with and without missing data.

In our analysis, the test returned a χ² statistic of 21.97 with 33 degrees of freedom and a p-value of 0.928. Since this p-value is well above the standard threshold of 0.05, we fail to reject the null hypothesis, indicating no significant differences between the observed data patterns. We therefore conclude that the data for the selected variables is MCAR. Furthermore, the Fisher’s Exact Test further confirms that there is no significant association between missingness in the subset of predictor variables and the outcome variable, pe_status, as indicated by a large p-value of 0.3221. This supports the conclusion that the missingness in this subset is independent of PE diagnosis and is therefore likely Missing Completely at Random (MCAR).

```{r}
vars_under_5 <- c("smoking_history", "emphysema_ct", "copd_history", "copd_exacerbation", 
                  "clinical_symptoms_dvt", "immobilization_surgery", "dvt_pe", 
                  "haemoptysis", "malignancy", "surgery_fracture", "active_malignancy", 
                  "unilateral_lower_limb_pain", "pain_lower_limb_edema", "well_score", 
                  "revised_geneva_score", "pe_more_likely")

data_subset<- merged_data[, vars_under_5]
mcar_test(data_subset)
```


```{r}
merged_data$any_missing_subset <- apply(is.na(data_subset), 1, any)
fisher.test(table(merged_data$any_missing_subset, merged_data$pe_status))
```

```{r}
merged_data <- merged_data %>%
  select(-c(any_missing_subset))
```

Because the missingness is determined to be completely at random (MCAR) and affects less than 5% of the data for each variable, we can apply either simple imputation or complete case analysis without introducing bias. We began by performing complete case analysis, removing all rows with missing values for the selected variables. This resulted in the exclusion of 9 rows, accounting for only 2.46% of the entire dataset. Such minimal data loss is unlikely to affect statistical power or validity, especially since 97.5% of the original sample is retained. 

```{r}
n_complete <- sum(complete.cases(data_subset))
n_total <- nrow(data_subset)
n_dropped <- n_total - n_complete

cat("Number of rows that would be dropped:", n_dropped, "\n")
cat("Percentage of data dropped:", round((n_dropped / n_total) * 100, 2), "%\n")
```

To ensure the robustness of our approach, we conducted a sensitivity analysis by comparing results from complete case analysis to those obtained using simple imputation. In the imputation method, missing values for categorical variables were replaced with the mode of each respective variable, while missing values for numerical variables were substituted with their corresponding mean. Logistic regression models were then fit on both the complete case and imputed datasets, and their AIC values were compared. The model based on complete case analysis yielded an AIC of 348.99, whereas the model based on the imputed dataset had an AIC of 353.70, suggesting that the complete case model provided a slightly better fit on the data that predicts the outcome variable, pe_status. Based on this comparison, we finalized our dataset using the complete case approach, resulting in the removal of 9 rows and retaining a total of 357 observations.

```{r}
data_complete_case <- na.omit(data_subset) 
get_mode <- function(x) {
  ux <- unique(x[!is.na(x)])
  ux[which.max(tabulate(match(x, ux)))]
}

data_imputed <- data_subset %>%
  mutate(across(where(is.numeric), ~ ifelse(is.na(.), mean(., na.rm = TRUE), .))) %>%
  mutate(across(where(is.factor), ~ ifelse(is.na(.), get_mode(.), .)))

data_complete_case$pe_status <- merged_data$pe_status[complete.cases(data_subset)]
data_imputed$pe_status <- merged_data$pe_status

model_cc <- glm(pe_status ~ ., data = data_complete_case, family = binomial)
model_imp <- glm(pe_status ~ ., data = data_imputed, family = binomial)

aic_values <- data.frame(
  Method = c("Complete Case Analysis", "Simple Imputation"),
  AIC = c(AIC(model_cc), AIC(model_imp))
)

aic_values %>%
  kable("html", caption = "Comparison of AIC Values: Complete Case vs Imputation") %>%
  kable_styling(full_width = FALSE, bootstrap_options = c("striped", "hover"))

merged_data <- merged_data[complete.cases(merged_data[, vars_under_5]), ]
```

**Handling Missingess of Variables with 5-20 percent missingness ** 

The only variable with moderate missingness (between 5–20%) is elevated_troponin, with 7.28% of values missing. To assess the nature of this missingness, we examine whether it is associated with the outcome variable, pe_status (pulmonary Embolism diagnosis).

A cross-tabulation shows that among individuals diagnosed with PE, 170 had observed elevated_troponin values while 17 had missing values. Among those without PE, 161 had observed values and 9 had missing values. The similar proportions of missingness in both groups suggest that the absence of data is not systematically linked to PE status—supporting the assumption that the data may be Missing Completely At Random (MCAR).

We also examined the PE rate within each group. Among those with observed troponin values, the PE rate is 48.6%, whereas it is 34.6% among those with missing troponin values. To determine whether this difference is statistically significant, we conducted both a Chi-square test of independence and Little’s MCAR test.

```{r}
troponin_pe_table <- table(is.na(merged_data$elevated_troponin), merged_data$pe_status)

troponin_pe_df <- as.data.frame.matrix(troponin_pe_table)
rownames(troponin_pe_df) <- c("Observed (Not Missing)", "Missing")
troponin_pe_df <- tibble::rownames_to_column(troponin_pe_df, var = "Troponin Missingness")
colnames(troponin_pe_df)[2:3] <- c("PE Absent (0)", "PE Present (1)")

troponin_pe_df %>%
  kable("html", caption = "Cross-tabulation of Troponin Missingness by PE Status") %>%
  kable_styling(bootstrap_options = c("striped", "hover", "condensed"), full_width = FALSE)
```

```{r}
merged_data %>%
  mutate(pe_status_num = as.numeric(as.character(pe_status)),
         troponin_missing = ifelse(is.na(elevated_troponin), "Missing", "Not Missing")) %>%
  group_by(troponin_missing) %>%
  summarise(pe_rate = mean(pe_status_num, na.rm = TRUE),
            n = n()) %>%
  kable("html", caption = "Pulmonary Embolism Rate by Elevated Troponin Missingness") %>%
  kable_styling(full_width = FALSE, bootstrap_options = c("striped", "hover"))
```

The Chi-square test yielded a p-value of 0.24, indicating no significant association between missingness in elevated_troponin and pe_status. The Little’s MCAR test also returned a non-significant p-value (0.168), reinforcing the result.

```{r}
chisq.test(table(is.na(merged_data$elevated_troponin), merged_data$pe_status))
mcar_test(merged_data[,c("elevated_troponin","pe_status")]) 
```

These findings suggest that the missingness in elevated_troponin is not systematically related to observed data, and thus can be considered MCAR. However, it is reasonable to proceed under the MAR (Missing At Random) framework for multiple imputation rather than complete case analysis. A complete case approach would result in the exclusion of 26 observations, which could reduce statistical power and introduce bias, especially if the missingness is related to the outcome variable, PE status. In order to handle the missing values in elevated_troponin, we applied Multiple Imputation by Chained Equations (MICE) using logistic regression imputation. 

The imputation model included the outcome variable pe_status and additional covariates to preserve the variability and account for potential relationships between elevated_troponin and other predictors. To identify relevant covariates for imputation, we trained a Random Forest classifier on the observed data and ranked the predictors based on their Mean Decrease in Accuracy (a measure of variable importance.) From variables with less than 20% missingness, we selected the top four predictors: acute_resp, map_value, bmi, and heart_rate. These were chosen based on their predictive strength. Additionally, we included age and pe_status (the outcome variable), clinical_symptoms_dvt (the most informative symptom-related feature), and systolic and diastolic blood pressure readings, as they are clinically relevant due to their association with elevated troponin and potential cardiac injury. These covariates, together with elevated_troponin, were included in the MICE procedure with 5 imputed datasets and 10 iterations, using the first completed dataset to impute the missing values.

```{r}
missing_prop <- colMeans(is.na(merged_data))
vars_under_20 <- names(missing_prop[missing_prop < 0.20]) 
```

```{r}
formula_str <- paste(
  "elevated_troponin ~", 
  paste(vars_under_20[vars_under_20 != "elevated_troponin"], collapse = " + ")
)
rf_formula <- as.formula(formula_str)

rf_data <- merged_data %>% filter(!is.na(elevated_troponin))

set.seed(123)
rf_model <- randomForest(
  formula = rf_formula,
  data = rf_data,
  importance = TRUE, 
  ntree = 500
)

var_importance <- importance(rf_model, type = 1) 
top_vars <- rownames(var_importance)[order(-var_importance[, 1])][1:10]

importance_df <- as.data.frame(importance(rf_model))
importance_df$Variable <- rownames(importance_df)

top10_vars <- importance_df %>%
  arrange(desc(MeanDecreaseAccuracy)) %>%
  select(Variable, MeanDecreaseAccuracy) %>%
  slice(1:10)

top10_vars %>%
  kable("html", caption = "Top 10 Predictors of Elevated Troponin (by Mean Decrease Accuracy)") %>%
  kable_styling(bootstrap_options = c("striped", "hover", "condensed"), full_width = FALSE)
```

```{r}
impute_vars <- c("pe_status", "elevated_troponin", "age", "bmi", "acute_resp", "map_value", "heart_rate")
impute_data <- merged_data[, impute_vars]
```

```{r, message=FALSE, warning=FALSE, results='hide'}
capture.output({
  imputed_data <- mice(
    impute_data,
    m = 5,
    method = "logreg",
    maxit = 10,
    seed = 123
  )
}, file = NULL)
```

```{r}
complete_data <- complete(imputed_data, 1)
merged_data$elevated_troponin <- complete_data$elevated_troponin
```

```{r}
pre_imputation_data <- merged_data %>%
  filter(!is.na(impute_data$elevated_troponin)) 
```

Sensitivity Analysis to Justify Multiple Imputation of elevated_troponin: The bar plot comparing observed and imputed values of elevated_troponin indicates that the proportions of levels (0 vs. 1) are reasonably consistent across both groups. This further suggests that the imputation preserved the original distribution of the data. 

To assess whether imputation influenced model results, we fit two logistic regression models: one using the complete-case data (before imputation) and another using the full dataset after imputing missing elevated_troponin values. Both models included the same set of relevant predictors used in the imputation model. The results show that the interpretation of the model remains consistent before and after imputation: The elevated_troponin variable is not a statistically significant predictor of pulmonary embolism in either model. In addition, the coefficient for elevated_troponin_1 changed insignificantly after imputation (-0.064 to -0.102) which indicates that the imputation of values did not significantly change the variable's relationship with the outcome variable, pe_status. The variables acute_resp and clinical_symptoms_dvt are statistically significant in both models. Other covariates remain non-significant, reinforcing the stability of model estimates. This supports the conclusion that multiple imputation did not introduce bias or meaningfully alter the relationships among variables. This validates the appropriateness of the imputation procedure.

```{r}
merged_data %>%
  mutate(trop_flag = ifelse(is.na(impute_data$elevated_troponin), "Imputed", "Observed")) %>%
  group_by(trop_flag, elevated_troponin) %>%
  summarise(n = n()) %>%
  mutate(prop = n / sum(n)) %>%
  ggplot(aes(x = elevated_troponin, y = prop, fill = trop_flag)) +
  geom_bar(stat = "identity", position = "dodge") +
  labs(title = "Proportion of Elevated Troponin (Observed vs Imputed)",
       x = "Elevated Troponin", y = "Proportion") +
  theme_minimal()
```

```{r}
impute_vars <- c("elevated_troponin", "age", "map_value", "bmi", "acute_resp", 
                 "heart_rate")

logit_formula <- as.formula(paste("pe_status ~", paste(impute_vars, collapse = " + ")))

model_complete <- glm(logit_formula,
                      data = pre_imputation_data,
                      family = binomial)

model_imputed <- glm(logit_formula,
                     data = merged_data,
                     family = binomial)

summary_complete <- summary(model_complete)
summary_imputed <- summary(model_imputed)

logit_table <- data.frame(
  Coef_Complete = round(summary_complete$coefficients[, "Estimate"], 3),
  P_Complete = round(summary_complete$coefficients[, "Pr(>|z|)"], 3),
  Coef_Imputed = round(summary_imputed$coefficients[, "Estimate"], 3),
  P_Imputed = round(summary_imputed$coefficients[, "Pr(>|z|)"], 3)
)


logit_table_sorted <- logit_table[order(logit_table$P_Imputed), ]

logit_table_sorted %>%
  kable("html", caption = "Logistic Regression Comparison Between Pre and Post Imputated Data") %>%
  kable_styling(bootstrap_options = c("striped", "hover", "condensed"), full_width = FALSE)

```

**Handling Missingness of Variables with more than 20 percent missingness ** 

```{r}
high_missing
```

The variables with more than 80 percent missingness include: rv_to_lv_ratio, rv_lv_abnormality, rv_edd_cm, and rv_edd_abnormality. The best approach to handle such variables is to remove them from the dataset. Both simple and multiple imputation methods are highly unreliable to deal with such a large proportion of missingness and might create biased results. After removing these variables, we end up with 48 remaining variables. 

```{r}
merged_data <- merged_data %>%
  select(-c(rv_to_lv_ratio, rv_lv_abnormality, rv_edd_cm, rv_edd_abnormality))
```

**Variables: echo_rvsp_value and rvsp_based_age ** 

Based on the cross-tabulation output, we observe that among individuals diagnosed with pulmonary embolism (PE), 82 had observed values for echo_rvsp_value (Echocardiographic RVSP Right Ventricular Systoic Pressure data), while 88 had missing entries (48 percent missing). In contrast, among those not diagnosed with PE, only 57 had observed values, and 130 had missing values (70 percent missing). This indicates that missingness in echo_rvsp_value is more prevalent among individuals without PE. The unequal distribution of missing values between the PE and non-PE groups suggests that the missingness is not completely at random (not MCAR), and may be associated with PE status. 

```{r}
echo_pe_table <- table(is.na(merged_data$echo_rvsp_value), merged_data$pe_status)
echo_pe_df <- as.data.frame.matrix(echo_pe_table)
rownames(echo_pe_df) <- c("Observed (Not Missing)", "Missing")
echo_pe_df <- tibble::rownames_to_column(echo_pe_df, var = "Echo RVSP Missingness")
colnames(echo_pe_df)[2:3] <- c("PE Absent (0)", "PE Present (1)")
echo_pe_df %>%
  kable("html", caption = "Cross-tabulation of Echo RVSP Missingness by PE Status") %>%
  kable_styling(bootstrap_options = c("striped", "hover", "condensed"), full_width = FALSE)
```

Furthermore, the chi-square test of independence yielded a small p-value (0.0008771), indicating a significant association between missingness in echo_rvsp_value and the outcome variable, pe_status. This further supports the conclusion that missingness in this variable is not completely at random (not MCAR). The Little's MCAR test also showcases a small p-value, confirming that missingess is not MCAR and that the chi-square test results are valid. 

```{r}
chisq.test(table(is.na(merged_data$echo_rvsp_value), merged_data$pe_status))
mcar_test(merged_data[,c("echo_rvsp_value","pe_status")]) 
```

To investigate the missingness mechanism of the echo_rvsp_value variable, we conducted a logistic regression analysis to determine whether the probability of missingness is associated with other observed variables. We first created a binary indicator variable, missing_echo_rvsp_value, where 1 indicates a missing value and 0 indicates an observed one. To avoid introducing bias or excessive row deletion—since the glm() function automatically excludes rows with any missing values—we restricted our analysis to a subset of variables that have no missing data. The resulting logistic regression model revealed that the variable pe_status was a significant predictor of missingness, suggesting that the missingness in echo_rvsp_value is not completely at random (MCAR) and is more likely Missing at Random (MAR), as it appears to depend on observed data.

```{r}
merged_data$missing_echo_rvsp_value <- as.numeric(is.na(merged_data$echo_rvsp_value))
```

```{r}
no_missing_vars <- names(merged_data)[colSums(is.na(merged_data)) == 0]
merged_data_no_missing <- merged_data[, no_missing_vars]
```

```{r}
excluded_vars <- c("echo_rvsp_value", "missing_echo_rvsp_value")
predictor_vars <- setdiff(names(merged_data_no_missing), excluded_vars)

formula_str <- paste("missing_echo_rvsp_value ~", paste(predictor_vars, collapse = " + "))
formula <- as.formula(formula_str)

missingness_model <- glm(formula, data = merged_data_no_missing, family = binomial)

significant_vars <- tidy(missingness_model) %>%
  filter(term != "(Intercept)", p.value < 0.1) %>%
  arrange(p.value)

significant_vars %>%
  mutate(p.value = signif(p.value, 3)) %>%
  kable("html", caption = "Significant Predictors of Missingness in echo_rvsp_value (p < 0.1)") %>%
  kable_styling(bootstrap_options = c("striped", "hover", "condensed"), full_width = FALSE)
```


In real-world clinical studies, simple and multiple imputation become unreliable when more than 50% of a variable’s data is missing, especially if the missingness is not completely random. As shown above, a greater proportion of patients with missing echo_rvsp_value were not diagnosed with pulmonary embolism, suggesting that the missingness may depend on unobserved factors—such as clinicians selectively ordering echocardiograms based on PE status. This implies that the missingness in the echocardiographic RVSP data is likely not MCAR and possibly not even MAR.

To address this, we will explore whether alternative markers of right ventricular strain can be used to complete the analysis. For instance, an echocardiogram is often used to estimate right ventricular systolic pressure (RVSP), which helps assess pulmonary hypertension which is another variable available in our dataset. However, even if the missingness in echo_rvsp_value is MAR (missingness depends on observed covariates), applying multiple imputation methods such as MICE on variables with over 50% missing data risks allowing the imputed values to dominate the dataset, potentially undermining the validity of the results. Therefore, we will exclude both echo_rvsp_value and its derived categorical/factor variable, rvsp_based_age, from the final analysis. 


**Variables: bnp_levels and elevated_bnp ** 

Based on the cross-tabulation output, among individuals diagnosed with pulmonary embolism (PE), 134 had observed values for bnp_levels (a numerical measure of B-type natriuretic peptide), while 53 had missing entries. Similarly, among individuals without PE, 123 had observed values and 47 had missing values. The comparable distribution of missingness between the PE and non-PE groups suggests that the missing values in bnp_levels are not systematically related to PE status. This pattern supports the assumption that the missingness may be Missing Completely at Random (MCAR).

```{r}
bnp_pe_table <- table(is.na(merged_data$bnp_levels), merged_data$pe_status)
bnp_pe_df <- as.data.frame.matrix(bnp_pe_table)
rownames(bnp_pe_df) <- c("Observed (Not Missing)", "Missing")
bnp_pe_df <- tibble::rownames_to_column(bnp_pe_df, var = "BNP Levels Missingness")
colnames(bnp_pe_df)[2:3] <- c("PE Absent (0)", "PE Present (1)")
bnp_pe_df %>%
  kable("html", caption = "Cross-tabulation of BNP Levels Missingness by PE Status") %>%
  kable_styling(bootstrap_options = c("striped", "hover", "condensed"), full_width = FALSE)
```

Furthermore, the chi-square test of independence yielded a small p-value (0.9776), indicating a nonsignificant association between missingness in bnp_levels and the outcome variable, pe_status. This further supports the conclusion that missingness in this variable is completely at random (MCAR). The Little's MCAR test also showcases a large p-value (p-value = 0.8838519), confirming that missingess is MCAR and that the chi-square test results are valid.

```{r}
chisq.test(table(is.na(merged_data$bnp_levels), merged_data$pe_status))
mcar_test(merged_data[,c("bnp_levels","pe_status")]) 
```

Logistic Regression: To investigate the missingness mechanism of the bnp_levels variable, we conducted a logistic regression analysis to assess whether the probability of missingness was associated with other observed variables. We first created a binary indicator variable, missing_bnp_levels, where 1 indicates a missing value and 0 indicates an observed one. To avoid introducing bias or excessive row deletion during logistic regression, we restricted the analysis to a subset of variables with no missing data. The logistic regression results indicated that pe_status was not a significant predictor of missingness. However, variables such as age and admission_year2020 weakly predicted missingness in bnp_levels. This suggests that while the missingness may appear close to Missing Completely at Random (MCAR), there is some dependence on observed variables, making the missingness mechanism more aligned with Missing at Random (MAR). Therefore, we proceed with multiple imputation rather than complete case analysis to retain statistical power and reduce bias.

```{r}
merged_data$missing_bnp_levels<- as.numeric(is.na(merged_data$bnp_levels))
```

```{r}
no_missing_vars <- names(merged_data)[colSums(is.na(merged_data)) == 0]
merged_data_no_missing <- merged_data[, no_missing_vars]
```

```{r}
excluded_vars <- c("bnp_levels", "missing_bnp_levels", "elevated_bnp")
predictor_vars <- setdiff(names(merged_data_no_missing), excluded_vars)

formula_str <- paste("missing_bnp_levels ~", paste(predictor_vars, collapse = " + "))
formula <- as.formula(formula_str)

missingness_model <- glm(formula, data = merged_data_no_missing, family = binomial)

significant_vars <- tidy(missingness_model) %>%
  filter(term != "(Intercept)", p.value < 0.1) %>%
  arrange(p.value)

significant_vars %>%
  mutate(p.value = signif(p.value, 3)) %>%
  kable("html", caption = "Significant Predictors of Missingness in bnp_levels (p < 0.1)") %>%
  kable_styling(bootstrap_options = c("striped", "hover", "condensed"), full_width = FALSE)
```

Imputation for bnp_levels: Missing values in bnp_levels were imputed using Predictive Mean Matching (PMM) via the MICE algorithm. The predictor variables used in the imputation include both: (1) variables most strongly associated with missingness in bnp_levels (as detailed in the table above), and (2) clinically relevant predictors known to be associated with both the outcome variable pe_status and with BNP (B-type Natriuretic Peptide), a key biomarker used to diagnose and assess heart failure. To address the skewed distribution of BNP, bnp_levels was log-transformed using log1p() prior to imputation. PMM was then applied over 5 imputed datasets and 10 iterations. After imputation, the values were back-transformed to their original scale using expm1(). Finally, the binary variable elevated_bnp was derived from the imputed bnp_levels, coded as 1 for values greater than 100 pg/mL, and 0 otherwise.

```{r}
mandatory_predictors <- c(
  "age", "admission_year", "copd_history", "pe_status", "bmi_marker"
)
clinical_predictors <- c(
  "heart_rate", "systolic", "diastolic"
)
impute_predictors <- c(mandatory_predictors, clinical_predictors)
```

```{r}
merged_data$log_bnp <- log1p(merged_data$bnp_levels)
impute_data <- merged_data[, c("log_bnp", impute_predictors)]
```

```{r}
pred_matrix <- make.predictorMatrix(impute_data)
pred_matrix[,] <- 0  # Start clean
pred_matrix["log_bnp", impute_predictors] <- 1
```

```{r, message=FALSE, warning=FALSE, results='hide'}
capture.output(
  imp <- mice(impute_data,
              method = "pmm", 
              predictorMatrix = pred_matrix,
              m = 5, 
              maxit = 10,
              seed = 123),
  file = NULL
)
```

```{r}
log_bnp_imputed <- complete(imp, 1)$log_bnp
merged_data$bnp_levels <- expm1(log_bnp_imputed)
```

```{r}
merged_data$bnp_levels <- as.numeric(merged_data$bnp_levels)
merged_data$elevated_bnp <- factor(
  ifelse(merged_data$bnp_levels > 100, 1, 0),
  levels = c(0, 1)
)
```

Validating Imputation of bnp_levels and elevated_bnp: The summary table presents descriptive statistics for bnp_levels by imputed_flag. The observed and imputed groups show similar central tendencies, with mean values of 262.84 (observed) and 251.37 (imputed), and median values of 70.99 and 66.01, respectively. As expected, the standard deviation is lower in the imputed group, likely due to both the smoothing effect of imputation and a smaller sample size. Additionally, results from the Welch two-sample t-test and Wilcoxon rank-sum test yield p-values greater than 0.05, indicating no statistically significant difference in the means or distributions of bnp_levels between the observed and imputed groups. This conclusion is further supported by the density plot, which shows substantial overlap between the two distributions and no notable shift post-imputation. Taken together, these findings confirm that the imputation of bnp_levels was successful and did not introduce meaningful bias.

```{r}
imputed_df <- complete(imp, 1)
imputed_df$imputed_flag <- ifelse(is.na(merged_data$log_bnp), "Imputed", "Observed")

imputed_df$bnp_levels <- expm1(imputed_df$log_bnp)

summary_table <- imputed_df %>%
  mutate(imputed_flag = factor(imputed_flag, levels = c("Observed", "Imputed"))) %>%
  group_by(imputed_flag) %>%
  summarise(
    Mean = round(mean(bnp_levels, na.rm = TRUE), 2),
    Median = round(median(bnp_levels, na.rm = TRUE), 2),
    SD = round(sd(bnp_levels, na.rm = TRUE), 2),
    N = n(),
    .groups = 'drop'
  )

summary_table %>%
  kable("html", caption = "Summary of BNP Levels by Imputation Status") %>%
  kable_styling(bootstrap_options = c("striped", "hover", "condensed", "responsive"))
```

```{r}
t.test(bnp_levels ~ imputed_flag, data = imputed_df)
wilcox.test(bnp_levels ~ imputed_flag, data = imputed_df)
```

```{r}
ggplot(imputed_df, aes(x = bnp_levels, fill = imputed_flag)) +
  geom_density(alpha = 0.4) +
  labs(
    title = "Distribution of Observed vs. Imputed BNP Levels",
    x = "BNP Levels",
    y = "Density",
    fill = "Value Type"
  ) +
  theme_minimal()
```

**Variables: posiive_d_dimer and d_dimer_value **

```{r}

```


**Variables: covid ** 


**Variables: dvt_in_lower_extremity ** 

**Variable: tricuspid_regurgitation ** 

**Variable: septal_bowing** 

**Variable: rh_strain ** 

**Variable: pulmonary_hypertension ** 

**Variable: rh_thrombus ** 

**Variable: rv_hypokines_on_echo **





# Part C: Descriptive Statistics (Univariate): Identifying Potential Outliers/Uncaptured Missing Values (No Stratification) 

**1. Categorical Variables ** 

Statistical Summaries (Counts, Frequencies, and Percentages) 

Visualizations: Bar Graphs, Pie Charts 

```{r}

```


**2. Continuous Variables (Numerical):** 

Statistical Summaries 

Visaulizations 

Identifying Outliers 


#Part D: Stratified Analysis (Bivariate): Stratify individuals by the presence of Pulmonary Embolism 




#Part E: Multivariate Analysis (Logistic Regression) 

First Research Question: 















